---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Published SHA-256 of the shipped rules database.
db_path <- file.path("inst", "db", "rules.db")
stopifnot(file.exists(db_path))
db_hash <- digest::digest(db_path, algo = "sha256", file = TRUE)

# Wrap in backticks for inline use.
backtick    <- intToUtf8(96L)
db_hash_md  <- paste0(backtick, db_hash, backtick)
```

# pkgaudit

[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
flags security-relevant files and code in R source packages for human review
without executing anything it scans. pkgaudit also models package lifecycle
execution semantics -- it reports not just what a package does, but when it
runs, so code that executes on install or load is distinguishable from code that
runs only when called.

A general-purpose scanner can read R -- Semgrep supports it experimentally --
but it reads files, not packages. Nothing tells it to look for R inside a help
file's `\examples{}` block or an `\Sexpr{}` macro, or inside a Sweave or R.rsp
vignette; and nothing tells it that `.onLoad()` runs on `library()`, that
`\Sexpr{}` evaluates while a help page is built, or that `configure` runs under
`R CMD check`. pkgaudit does this: it extracts R from wherever a package carries
it, parses the code for security-relevant patterns, and reports the lifecycle
phases in which each finding runs.

## Overview

pkgaudit identifies every file in a package that can contain code -- complete
source files (e.g., `R/*.R`, `src/*.c`), and code carried inside documentation
(`man/*.Rd`) and vignettes (e.g., `vignettes/*.Rmd`) -- and says what it made of
each one:

- **parsed**: R is parsed for security-relevant patterns like `system()` and
  `httr::POST()` calls. This includes `.R` scripts in `R/`, `data/`, `demo/`,
  `exec/`, `tests/`, and `tools/`; examples and `\Sexpr{}` macros in `.Rd`
  files; R chunks in R Markdown, Quarto, Sweave, and R.rsp vignettes; and
  `inst/CITATION`, `.Rprofile`, and `src/install.libs.R`.
- **matched**: Shell scripts and Make-like files, if present, are matched
  against regular expressions for security-relevant commands like `curl`. This
  includes `configure`, `cleanup`, `src/Makevars`, and more. Matching is
  textual, so a `curl` in a comment reads the same as one in command position.
- **exportable**: C, C++, Fortran, Rust, Python, and JavaScript source files,
  and vignette chunks written in those languages, are not read, but
  `export_unscanned()` can write them out for tools like Semgrep, blank-padded
  so that a finding's line number still points into the original file.
- **unexamined**: Files that may contain code but were not scanned, such as
  serialized `.rda` and `.rds` files, are reported as unexamined.
- **error**: If pkgaudit attempted to parse or grep a file but encountered an
  error, the file is reported as an error.

Coverage may not be complete. What pkgaudit offers is not completeness but
legibility: a clean result can be checked rather than trusted, because the scan
says what it did not cover.

Every finding carries the R package lifecycle phases in which it runs --
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`at_load`, `at_attach`, `at_unload`, `at_detach`, or `none` -- so code that
executes on `library()` is distinguishable from code that runs only when someone
calls it.

A finding is not an accusation. `configure` scripts and calls to system tools,
for example, are often legitimate. pkgaudit helps to identify what deserves
reviewer attention, not what is malicious.

For why this matters, see
[R Package Security](https://tylerjssmith.github.io/pkgaudit/articles/security.html).
For the rule set, see
[Rule Coverage](https://tylerjssmith.github.io/pkgaudit/articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Four of `untrustedpkg`'s five findings run without anyone asking for them.
`.onLoad()` in `R/zzz.R` calls `system()`, so it runs on `library()` -- and at
build, check, and source installation, each of which loads the package. An
`\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help page
is rendered, at those same three phases, and the `\examples{}` block in the same
file calls `download.file()`, which `R CMD check` runs. The `configure` script
may invoke `curl` at build, check, and source installation. The fifth finding is
the contrast that makes the rest legible: `download.file()` in `R/fetch.R` sits
in an ordinary function body, so it runs at no phase at all, reported under
`none`, because it executes only if someone calls it.

Phases overlap -- building a package with vignettes also installs and loads it
-- so one occurrence is counted under every phase it runs in. `path = FALSE`
omits the local filesystem path from output that will be shared.

Two functions carry a scan into other tools. `emit_sarif()` renders the result
as SARIF 2.1.0, which editors and code-scanning platforms read directly.
`export_unscanned()` writes the code pkgaudit cannot read into a directory for
a scanner that can. Both are covered in
[Getting Started with pkgaudit](https://tylerjssmith.github.io/pkgaudit/articles/pkgaudit.html).

## Rule Database Integrity

pkgaudit detects security-relevant files and code using a SQLite database of
rules shipped at `inst/db/rules.db`. `load_rules()` verifies the database
against its bundled `.sha256` sidecar on every call and refuses to load a
modified one. To check an installed copy against the value published here:

``` r
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

Expected SHA-256: `r db_hash_md`

## Security

pkgaudit's own security model, and how to report a vulnerability in it, are in
[SECURITY.md](.github/SECURITY.md). To propose or revise a rule, see
[CONTRIBUTING.md](.github/CONTRIBUTING.md).
