---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Published SHA-256 of the shipped rules database.
db_path <- file.path("inst", "db", "rules.db")
stopifnot(file.exists(db_path))
db_hash <- digest::digest(db_path, algo = "sha256", file = TRUE)

# Wrap in backticks for inline use.
backtick    <- intToUtf8(96L)
db_hash_md  <- paste0(backtick, db_hash, backtick)
```

# pkgaudit

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit scans R packages for security-relevant files and code without executing
anything it scans. It reports what code does and when it runs, so code that runs
automatically on install or load is distinguishable from code that runs only
when called.

A general-purpose scanner like Semgrep can read R -- but it reads scripts, not
packages. It does not look for R inside an `\examples{}` block, an `\Sexpr{}`
macro, or a vignette. It does not know that `.onLoad()` runs when a user
calls `library()` or that a `configure` script runs on `R CMD check`. pkgaudit
extracts code wherever it exists, scans it for functions and commands that need
human review, and reports the lifecycle phases in which each finding runs.

For why this matters, see
[R Package Security](https://tylerjssmith.github.io/pkgaudit/articles/security.html).
For the rule set, see
[Rule Coverage](https://tylerjssmith.github.io/pkgaudit/articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration. We see that, when `untrustedpkg` is installed from source, its R
code will make an HTTP request (`httr`) and invoke a shell command (`system`).
Meanwhile, a shell script or Make-like file make invoke the `curl` command.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, phase = "at_install_src", path = FALSE)
```

The supported phases are: `at_autoconf`, `at_build`, `at_check`,
`at_install_src`, `at_install_bin`, `at_load`, `at_attach`, `at_unload`, and
`at_detach`. A finding with no phase is reported as `none`.

pkgaudit can integrate its scan with other tools. `emit_sarif()` renders its
results as SARIF 2.1.0, which editors and code-scanning platforms read directly.
`export_unscanned()` exports code written in languages pkgaudit cannot read to a
directory for a scanner that can. 

A finding is not an accusation. `configure` scripts and calls to system tools,
for example, are often legitimate. pkgaudit helps to identify what deserves
reviewer attention, not what is malicious.

See
[Getting Started with pkgaudit](https://tylerjssmith.github.io/pkgaudit/articles/pkgaudit.html)
for details.

## Rule Database Integrity

pkgaudit detects security-relevant files and code using a SQLite database of
rules shipped at `inst/db/rules.db`. `load_rules()` verifies the database
against its bundled `.sha256` sidecar on every call and refuses to load a
modified one. To check an installed copy against the value published here:

``` r
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

Expected SHA-256: `r db_hash_md`

## Security

pkgaudit's own security model, and how to report a vulnerability, are in
[SECURITY.md](.github/SECURITY.md). To propose or revise a rule, see
[CONTRIBUTING.md](.github/CONTRIBUTING.md). How pkgaudit works internally, for a
reader auditing the source, is in
[Internals](https://tylerjssmith.github.io/pkgaudit/articles/internals.html).

## Citation

```{r citation, eval = FALSE}
citation("pkgaudit")
```
