---
output:
  github_document:
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r load, message = FALSE}
library(tidyverse)
library(pkgaudit)

source("scripts/download_cran.R")
source("scripts/survey_cran.R")
source("scripts/survey_tarballs.R")

SNAPSHOT_DIR = "../../../snapshot"
```

# Survey of R Packages on CRAN

## Download CRAN

The following code obtain a list of available packages.

```{r available, eval = FALSE}
available = available.packages("https://cran.rstudio.com/src/contrib")
available = as.data.frame(available[, c("Package","Version","MD5sum")])
available = available[!duplicated(available), ]
write_csv(available, "results/available_packages.csv")
```

The following code loads the list of available packages we obtained.

```{r available_saved}
available_saved = read_csv("results/available_packages.csv",
  show_col_types = FALSE)
head(available_saved)
```

The following code downloads the available packages as source tarballs. It
confirms that the MD5 hash claimed by the mirror matches the MD5 hash of the
downloaded file, and adds the more secure (collision-resistant) SHA-256 hash.

The snapshot is large (~18 GB) and was saved outside of the package directory at
the path stored in `SNAPSHOT_DIR`.

```{r download, eval = FALSE, message = FALSE}
downloaded = download_cran(
  pkgs       = available[sample(1:nrow(available), 20), ], 
  dir        = SNAPSHOT_DIR, 
  check_md5  = TRUE, 
  add_sha256 = TRUE,
  pause = 0
)
write_csv(downloaded, "results/downloaded_packages.csv")
```

The following code loads the list of downloaded packages.

```{r download_saved}
downloaded_saved = read_csv("results/downloaded_packages.csv",
  show_col_types = FALSE)
head(downloaded_saved[, c("package","version","md5_matches","sha256_local")])
```

## Survey CRAN

The following code runs `pkgaudit::audit_tarball` on every downloaded package,
combines the `pkgaudit` data frames across packages, and adds `package` and
`version` columns to distinguish the results.

```{r cran, eval = FALSE}
findings = survey_cran(SNAPSHOT_DIR, rules = pkgaudit::load_rules())
write_rds(findings, "results/findings.rds", compress = "gz")
```

The following code loads the survey results, extracts the data frames, defines
the total number of tarballs surveyed as the denominator, and defines two
functions used to summarize the results.

```{r setup_results}
# Load Saved Results
saved_findings = read_rds("results/findings.rds")

file_contexts = saved_findings$file_contexts
patterns      = saved_findings$patterns
matches       = saved_findings$matches
coverage      = saved_findings$coverage
errors        = saved_findings$errors
provenance    = saved_findings$provenance

# Get Denominator
N = length(list.files(SNAPSHOT_DIR, pattern = "\\.tar\\.gz$"))

# Define Summary Functions
get_proportion_hits <- function(data, x) {
  data %>%
    count({{ x }}) %>%
    mutate(p = n / sum(n) * 100) %>%
    arrange(desc(n))
}

get_proportion_pkgs <- function(data, x) {
  data %>%
    group_by({{ x }}) %>%
    summarise(n = n_distinct(package)) %>%
    mutate(p = n / N * 100) %>%
    rename(!!sym(paste0("n/", N)) := p) %>%
    arrange(desc(n))
}
```

### File Contexts

```{r file_contexts}
file_contexts %>%
  get_proportion_hits(file_context)
file_contexts %>%
  get_proportion_pkgs(file_context)
```

### Shell/Make-like Matches

```{r matches}
matches %>%
  get_proportion_hits(rule)
matches %>%
  get_proportion_pkgs(rule)
```

### R Patterns

```{r patterns}
patterns %>%
  get_proportion_hits(rule)
patterns %>%
  get_proportion_pkgs(rule)
```

## Tarballs

The following code runs `survey_structure()` and `survey_expansion()` on every
downloaded package.

```{r tarball, eval = FALSE}
structure <- survey_dir(SNAPSHOT_DIR, survey_structure)
expansion <- survey_dir(SNAPSHOT_DIR, survey_expansion)
```

The defaults `max_entries = 100000L` and `max_bytes = 2 * 1024^3` (2 GB) provide
considerable room above the empirical maxima.

```{r}
# max_entries
quantile(structure$n_entries,  c(.5, .9, .99, .999, 1))

# max_bytes
quantile(structure$bytes_read, c(.5, .9, .99, .999, 1))
```

The default `max_ratio = 256` likewise provides considerable room above the
empirical maximum. 

```{r}
# exclude capped and errored rows, if present, as ratios are lower bounds
ok <- subset(expansion, !cap_hit & is.na(error))
round(nrow(ok) / nrow(expansion) * 100, 4)

# max_ratio
quantile(ok$ratio, c(.5, .9, .99, .999, 1))
```

We can confirm the absence of various threat indicators in CRAN packages such
that the presence of any one justifies failing closed.

```{r}
sum(structure$n_symlink > 0)
sum(structure$n_hardlink > 0)
sum(structure$n_other_type > 0)
sum(structure$n_traversal > 0)
sum(structure$n_absolute > 0)
table(structure$n_toplevel)
sum(structure$n_backslash > 0)
sum(structure$n_control > 0)
sum(structure$n_empty > 0)
sum(structure$n_bad_size > 0)
```

```{r}
subset(structure, truncated | !is.na(error))
subset(expansion, cap_hit | !is.na(error))
```


