---
title: "pkgaudit Internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pkgaudit Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = FALSE
)
library(pkgaudit)
```

This vignette is for someone reading pkgaudit's source before trusting it. It
describes how a scan works, states the invariants the design rests on, and says
where in the source each one is enforced, so that the claims can be checked
rather than taken on trust.

For what pkgaudit reports, see
[Getting Started](pkgaudit.html); for the rule set, see
[Rule Coverage](rules.html). pkgaudit's own security model and its vulnerability
reporting policy are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).

## A scan in four movements

`audit_package()` reads top to bottom as a sequence of named concerns. It names
no file format and no language: those are decided by dispatch.

1. **Discover.** `find_file_contexts()` runs one `list.files()` per
   file-context rule, using the rule's `path`, `filename` and `recursive`
   fields. Every rule tells the scan which files to read; only a rule with
   `report = TRUE` contributes a row to `file_contexts`.
2. **Extract.** `extract_segments()` reads each file into *segments*: contiguous
   runs of code in one language, blank-padded to the length of the source so
   that line *n* of a segment is line *n* of the file. That padding is why a
   finding's line and column point into the original file with no offset to
   apply.
3. **Analyse.** `analyze_segment()` finds code contexts, patterns and matches in
   each segment.
4. **Resolve.** Lifecycle phases are attached once, at the end.

The findings frames are built **without** phase columns throughout and joined
once in step 4. Nothing before it knows about phases.

A phase is a property of two things -- where the file sits in the package, and
where the code sits within that file -- so step 4 resolves a pattern by walking
a short chain: a code context carrying phases of its own wins; otherwise the
code takes the phases of whatever encloses it. Code inside a function
definition takes them too, unless the file's rule sets `assume_called: FALSE`,
in which case it carries none. Only the rules for `R/` do.

That the frames reach step 4 knowing where they came from is why `patterns`
carries two extra columns while a scan is running, `file_rule` and
`rd_context`. `audit_package()` drops them before building the result, so
the object a caller receives has the columns `?audit_package` documents and no
others.

## Two axes of dispatch

Steps 2 and 3 are both S3 generics, and they dispatch on different things.

- `extract_segments()` dispatches on the file-context rule's **`type`** -- how
  the file is read. Methods live one per file in `R/extract_*.R`.
- `analyze_segment()` dispatches on the segment's **`language`** -- how the code
  is analysed. Methods live in `R/analyze_*.R`.

They cannot be one axis, because one file can yield code in more than one
language: an `.Rmd` yields an R segment for each `{r}` chunk and a shell segment
for each `{bash}` chunk. An `.Rd` yields several segments that are all R.

A type read exactly like another inherits from it rather than repeating a
method: `make` inherits `shell`, and `qmd` inherits `Rmd`.

Both defaults fail closed. A type with no reader yields no segments;
a language with no analyser yields no findings and no error, and records a
coverage row instead, so an unhandled chunk engine is accounted for rather than
silently dropped.

Adding a file format is one new `R/extract_*.R` plus a rule. Adding a language is
one new `R/analyze_*.R` plus rules carrying that language. Neither touches
`audit_package()`, and neither touches the other.

## The call graph

The graph below is derived from pkgaudit's own source, by parsing it with the
same parser the scan uses. It is not maintained by hand and cannot fall behind
the code.

```{r call-graph, comment = ""}
# Read the package's own sources. Present both in the source tree and in a
# built tarball, so this resolves wherever the vignette is built from.
r_dir <- if (dir.exists("../R")) "../R" else "R"
stopifnot(dir.exists(r_dir))

# Every function the package defines, and which of them each one calls. The
# names in function position are all.names() less all.vars(), which is the
# standard approximation and needs no dependency.
defined <- list()
for (f in list.files(r_dir, pattern = "[.]R$", full.names = TRUE)) {
  for (e in parse(f, keep.source = FALSE)) {
    if (!is.call(e) || length(e) < 3L) next
    if (!as.character(e[[1L]]) %in% c("<-", "=")) next
    if (!is.call(e[[3L]]) || !identical(as.character(e[[3L]][[1L]]), "function")) next
    nm <- as.character(e[[2L]])
    defined[[nm]] <- setdiff(all.names(e[[3L]]), all.vars(e[[3L]]))
  }
}
own <- names(defined)

# S3 dispatch is invisible to a static reader: a generic calls UseMethod(), not
# its methods. The methods are attached to their generic so the tree shows what
# a scan actually does.
for (g in c("extract_segments", "analyze_segment")) {
  methods <- grep(paste0("^", g, "[.]"), own, value = TRUE)
  defined[[g]] <- unique(c(setdiff(defined[[g]], "UseMethod"), methods))
}

render <- function(fn, prefix = "", seen = character(0), depth = 0L) {
  callees <- intersect(defined[[fn]], own)
  callees <- setdiff(callees, seen)
  if (depth >= 4L) callees <- character(0)
  out <- character(0)
  for (i in seq_along(callees)) {
    last <- i == length(callees)
    out <- c(out, paste0(prefix, if (last) "└─ " else "├─ ",
                         callees[[i]], "()"),
             render(callees[[i]], paste0(prefix, if (last) "   " else "│  "),
                    c(seen, callees), depth + 1L))
  }
  out
}

cat("audit_package()", render("audit_package", seen = "audit_package"),
    sep = "\n")
```

A name appears once, at the first place it is reached; a function called from
several places is not repeated. The tree is cut at four levels deep.

## Invariants

Each of these is a property a contributor could otherwise break, and each is
asserted by a test rather than left to review.

### Nothing that is scanned is ever executed

This is not a property any single function can be inspected for. R's own help
machinery will evaluate `\Sexpr` if asked, and `tools::prepare_Rd()`,
`Rd2HTML()` and their kin do exactly that. It is therefore asserted end to end:
`tests/testthat/test-no_execution.R` builds a package whose every execution site
writes a marker file, scans it, exports it, and requires that no marker exists
afterwards.

Three calls in the read path look like execution and are not.
`tools::parse_Rd()` parses without evaluating; `tools::loadPkgRdMacros()` reads
macro definitions as text; and `parse(text = )` produces an unevaluated
expression. A macro provider is looked up by macro name against a fixed list and
never taken from the audited package's `RdMacros` field, which the package under
audit chooses.

### The size limit runs before dispatch

`extract_segments()` applies the scanning limit before `UseMethod()`, so no
reader can be written that skips it. A per-method guard is exactly what once let
help files bypass the limit for a whole release.

### Failures are contained, and counted

Listing files, reading a file, parsing code, and evaluating an XPath or a
regular expression are each wrapped: a failure becomes a row in `errors` rather
than aborting the scan. The `Errors` section of `summary()` follows the table
with one note per step, stating what coverage that step's failures cost.

A file pkgaudit tried to read and could not is `error` in the coverage frame,
which is a different claim from `unexamined`, where it never tried. Only a
failure at a *reading* step counts: a rule that fails on a file says nothing
about the file.

### The rules are data, and integrity-checked

Rules live in a SQLite database built from YAML by `inst/scripts/build_db.R`.
`load_rules()` verifies the database against its SHA-256 sidecar on every call.
The check is time-of-check to time-of-use; see "Security considerations" in
`?load_rules`.

`load_rules()` also refuses a database with a gap in its phase model: a rule
without phases, a file-context rule naming a code context that does not exist,
or one that leaves `assume_called` undecided where code contexts can arise --
or decides it where they cannot. A rule can never ship whose findings would
silently carry no phases.

### Every rule is verified against its own examples

Each rule carries positive and negative examples in its YAML. `build_db.R`
writes them out as fixtures, and `tests/testthat/test-fixtures.R` requires every
positive to match its rule and every negative not to. A rule is trusted because
its documentation is executable, not because it is documented.

For a code-context rule matched on a segment label rather than on the parse
tree, the examples are help files and the test is a round trip: the extractor
must stamp the label the rule claims, on a page carrying that kind of code and
on no other. That is the only thing tying such a rule to the extractor, since
its label is checked against a fixed vocabulary when the database is built but
nothing there confirms the extractor produces it.

The same file checks two further contracts that would otherwise drift. A pattern
rule's `functions` field -- the names it claims for `do.call("name", ...)` -- is
accepted only if calling that name bare would have matched the rule's own XPath,
so an indirect finding can never be attributed to a rule that would not have
reported the direct call. And `eval_parse`'s source list is required to cover
every function the capability rules it composes declare.

Documentation is held to the same standard. The rule set's vignette is generated
from the shipped database rather than written alongside it, and every rule in
that database must be described there, so a rule that ships without a
description fails the vignette build. Neither that nor a failing example can be
fixed by editing prose.

## What pkgaudit does not do

- It does not execute, install, build, or load the package under audit.
- It does not resolve values. A path assembled at runtime, a name built by
  `paste0()`, or a chunk option computed at render time is not followed.
- It does not read compiled code, or any language other than R and shell.
  `export_unscanned()` hands those to a tool that does.
- It does not rank findings or return a verdict. There is no severity model,
  and the `level` in SARIF output is a mapping of the phase model, not a
  judgement.
- It does not reach the network, and it writes only where a caller sends it.
- It does not trace call graphs. Whether a particular package calls a particular
  function is not something pkgaudit determines, which is why the phases of code
  inside a function definition are a stated reading rather than a finding.
