---
title: "pkgaudit Internals"
output:
  rmarkdown::html_vignette:
    # Rule messages quote shell flags and make constructs verbatim, so the
    # "smart" typography that would turn --clean into an en dash, or
    # $(shell ...) into an ellipsis, is disabled.
    md_extensions: "-smart"
vignette: >
  %\VignetteIndexEntry{pkgaudit Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = FALSE
)

# Render a markdown table, one row per line, escaping any cell content that
# would otherwise be read as a column separator.
emit_table <- function(headers, ...) {
  cells <- lapply(list(...), function(x) gsub("|", "\\|", x, fixed = TRUE))
  cat(
    c(
      paste0("| ", paste(headers, collapse = " | "), " |"),
      paste0("|", strrep("---|", length(headers))),
      paste0("| ", do.call(paste, c(cells, sep = " | ")), " |")
    ),
    sep = "\n"
  )
  cat("\n\n")
}

library(pkgaudit)
```

This vignette is for someone reading pkgaudit's source code. For how to use
pkgaudit and what it reports, see [Getting Started](pkgaudit.html); for the rule
set, see [Rule Coverage](rules.html).

## pkgaudit steps

`audit_package()` orchestrates a series of function calls, which may be loosely
grouped as four steps:

1. Discover
2. Extract
3. Analyze
4. Resolve

### Discover

`find_file_contexts()` runs one `list.files()` per file-context rule, using a
file context rule's `path`, `filename` and `recursive` fields. If
`report = TRUE` in a rule, the file context is reported in the pkgaudit
object's `$file_contexts` data frame.

### Extract

For each file context, whether reported or not, `extract_segments()` dispatches
to S3 methods that read each file into *segments*: contiguous runs of code in
one language, blank-padded so that line *n* of a segment is line *n* of the
source file. Dispatch is based on the file-context rule's `type`. The methods
live in `R/extract_*.R`.

```{r results = "asis"}
emit_table(
  "Extractors",
  list.files(
    path    = "../R",
    pattern = "^extract_[a-zA-Z]+\\.R$"
  )
)
```

A type read like another inherits from it rather than repeating a method. For
example, `make` inherits `shell` and `qmd` inherits `Rmd`. `extract_segments()`
has a default method that fails closed: a type with no extractor yields no
segments.

### Analyze

For each segment extracted, `analyze_segment()` dispatches to S3 methods that
call functions to find code contexts, patterns, or matches in each segment.
Dispatch is based on the segment's `language`, such as R or shell. The methods
live in `R/analyze_*.R`.

```{r results = "asis"}
emit_table(
  "Analyzers",
  list.files(
    path    = "../R",
    pattern = "^analyze_[a-zA-Z]+\\.R$"
  )
)
```

`analyze_segment.R` (in `R/analyze_R.R`) calls functions including
`parse_code()`, which uses `parse()` and `xmlparsedata::xml_parse_data()` to
derive an XML parse tree for R code; and `find_patterns()` and
`find_indirect()`, which use the parse tree and `xml2::xml_find_all()` to find
patterns and indirect calls using function names (e.g.,
`do.call("system", ...)`). `determine_code_contexts()` determines the code
context of patterns and indirect calls.

`analyze_segment.shell` (in `R/analyze_shell.R`) calls functions including
`find_matches()`, which uses `gregexpr(perl = TRUE)` to find text matching
regular expressions in shell and Make-like files, and `bash`, `sh`, and `zsh`
vignette chunks.

`analyze_segment()` has a default method that fails closed: a language with no
analyzer yields no findings and no error, recording a coverage row instead.

### Resolve

A set of internal functions uses file or code contexts to resolve the phases in
which a pattern or match may run. The supported phases are:

```{r phases-legend, results = "asis"}
rules       <- load_rules()
phase_names <- setdiff(names(rules$phases), c("context", "version"))

phase_descriptions <- c(
  at_autoconf    = "Autoconf is run to generate `configure`",
  at_build       = "`R CMD build`",
  at_check       = "`R CMD check`",
  at_install_src = "`R CMD INSTALL` from source",
  at_install_bin = "a prebuilt binary package is installed",
  at_load        = "the namespace is loaded",
  at_attach      = "the package is attached to the search path",
  at_unload      = "the namespace is unloaded",
  at_detach      = "the package is detached from the search path"
)
stopifnot(setequal(names(phase_descriptions), phase_names))

emit_table(
  c("Phase", "Code executes when"),
  paste0("`", phase_names, "`"),
  phase_descriptions[phase_names]
)
```

Pattern phases are resolved from code and file contexts, in that order. A
pattern in a code context defined by a rule, such as `.onLoad`, takes that
context's phases. A pattern in top-level code takes its file context's phases.
A pattern in a regular function definition depends on the `assume_called` field
of the file-context rule: if `TRUE`, it takes the phases of whatever encloses
it; if `FALSE`, as under `R/`, it receives no phase. (A helper defined under
`R/` may still be called from a lifecycle hook, and human reviewers should
verify whether that occurs.) Match phases are resolved from file contexts. A
rule can belong to several phases.

Phases were established by running instrumented packages and recording which
sites fired, rather than read from documentation.

## Call list

The `audit_package()` call list below is derived from pkgaudit's source code.
A name appears once, at the first place it is reached; a function called from
several places is not repeated. The tree is cut at four levels deep.

```{r call-graph, comment = ""}
# Read the package's own sources.
r_dir <- if (dir.exists("../R")) "../R" else "R"
stopifnot(dir.exists(r_dir))

# Every function the package defines, and which of them each one calls. The
# names in function position are all.names() less all.vars().
defined <- list()
for (f in list.files(r_dir, pattern = "[.]R$", full.names = TRUE)) {
  for (e in parse(f, keep.source = FALSE)) {
    if (!is.call(e) || length(e) < 3L) next
    if (!as.character(e[[1L]]) %in% c("<-", "=")) next
    if (!is.call(e[[3L]]) || !identical(as.character(e[[3L]][[1L]]), "function")) next
    nm <- as.character(e[[2L]])
    defined[[nm]] <- setdiff(all.names(e[[3L]]), all.vars(e[[3L]]))
  }
}
own <- names(defined)

# S3 dispatch is invisible to a static reader: a generic calls UseMethod(), not
# its methods. The methods are attached to their generic so the tree shows what
# a scan actually does.
for (g in c("extract_segments", "analyze_segment")) {
  methods <- grep(paste0("^", g, "[.]"), own, value = TRUE)
  defined[[g]] <- unique(c(setdiff(defined[[g]], "UseMethod"), methods))
}

render <- function(fn, prefix = "", seen = character(0), depth = 0L) {
  callees <- intersect(defined[[fn]], own)
  callees <- setdiff(callees, seen)
  if (depth >= 4L) callees <- character(0)
  out <- character(0)
  for (i in seq_along(callees)) {
    last <- i == length(callees)
    out <- c(out, paste0(prefix, if (last) "└─ " else "├─ ",
                         callees[[i]], "()"),
             render(callees[[i]], paste0(prefix, if (last) "   " else "│  "),
                    c(seen, callees), depth + 1L))
  }
  out
}

cat("audit_package()", render("audit_package", seen = "audit_package"),
    sep = "\n")
```

## Invariants

### Scanned files are never executed

Since pkgaudit is intended for reviewing untrusted packages, it must never
execute code contained in those packages. `tests/testthat/test-no_execution.R`
builds a package where every execution site writes a marker file, scans it,
exports it, and requires that none of the markers exist afterwards.

Three calls in the read path look like execution and are not.
`tools::parse_Rd()` parses without evaluating; `tools::loadPkgRdMacros()` reads
macro definitions as text; and `parse(text = )` produces an unevaluated
expression. Macros come from the package's own `man/macros/` and from any
provider named in its `RdMacros` field; both are read as text, not evaluated.

### File size limits are applied before extraction

A file over 10 MB is not read. `extract_segments()` applies this restriction
before `UseMethod()` so that no method can skip it.
`tests/testthat/test-segments.R` verifies that an oversized file is not scanned
and that the refusal is recorded in `errors`.

### Failures are contained and counted

Listing files, reading a file, parsing code, and evaluating an XPath or a
regular expression are each wrapped in `tryCatch()`: a failure becomes a row in
`errors` rather than aborting the scan.

### Rules are integrity-checked

Rules live in a SQLite database built from YAML by `inst/scripts/build_db.R`.
`load_rules()` verifies the database against its SHA-256 sidecar on every call.
The check is time-of-check to time-of-use; see "Security considerations" in
`?load_rules`.

`load_rules()` also refuses a database with a gap in its phase model, so a rule
whose findings would carry no phases can never ship.

### Every rule is verified against its own examples

Each rule carries positive and negative examples in its YAML. `build_db.R`
writes them out as fixtures, and `tests/testthat/test-fixtures.R` requires every
positive to match its rule and every negative to not match.
