---
title: "pkgaudit Internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pkgaudit Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = FALSE
)
library(pkgaudit)
```

This vignette is for someone reading pkgaudit's source before trusting it. For
how to use pkgaudit and what it reports, see
[Getting Started](pkgaudit.html); for the rule set, see
[Rule Coverage](rules.html).

## pkgaudit steps

`audit_package()` orchestrates a series of function calls, which may be grouped
as follows:

1. **Discover.** `find_file_contexts()` runs one `list.files()` per
   file-context rule, using a file context rule's `path`, `filename` and
   `recursive` fields. If `report = TRUE` in a rule, the file context is
   reported in the pkgaudit object's `$file_contexts` data frame.
2. **Extract.** `extract_segments()` dispatches to S3 methods that read each
   file into *segments*: contiguous runs of code in one language, blank-padded
   so that line *n* of a segment is line *n* of the source file.
3. **Analyze.** `analyze_segment()` dispatches to S3 methods that call functions
   to find code contexts, patterns, or matches in each segment.
4. **Resolve.** A set of functions attaches and resolves the phases in which a
   pattern or match may run.

## Extraction and analysis

Steps 2 and 3 dispatch to S3 methods.

- `extract_segments()` dispatches on the file-context rule's **`type`**, such as
   an R source file, an Rd file containing examples, or an Rmd vignette
   containing code chunks. These methods live in `R/extract_*.R`.
- `analyze_segment()` dispatches on the segment's **`language`**, such as R or
   shell. These methods live in `R/analyze_*.R`.

A type or language read like another inherits from it rather than repeating a
method. For example, `make` inherits `shell`, and `qmd` inherits `Rmd`.

The default methods fail closed. A type with no extractor yields no segments; a
language with no analyzer yields no findings and no error, recording a coverage
row instead.

## Call list

The `audit_package()` call list below is derived from pkgaudit's source code. It
is not maintained by hand and cannot fall behind the code. A name appears once,
at the first place it is reached; a function called from several places is not
repeated. The tree is cut at four levels deep.

```{r call-graph, comment = ""}
# Read the package's own sources.
r_dir <- if (dir.exists("../R")) "../R" else "R"
stopifnot(dir.exists(r_dir))

# Every function the package defines, and which of them each one calls. The
# names in function position are all.names() less all.vars().
defined <- list()
for (f in list.files(r_dir, pattern = "[.]R$", full.names = TRUE)) {
  for (e in parse(f, keep.source = FALSE)) {
    if (!is.call(e) || length(e) < 3L) next
    if (!as.character(e[[1L]]) %in% c("<-", "=")) next
    if (!is.call(e[[3L]]) || !identical(as.character(e[[3L]][[1L]]), "function")) next
    nm <- as.character(e[[2L]])
    defined[[nm]] <- setdiff(all.names(e[[3L]]), all.vars(e[[3L]]))
  }
}
own <- names(defined)

# S3 dispatch is invisible to a static reader: a generic calls UseMethod(), not
# its methods. The methods are attached to their generic so the tree shows what
# a scan actually does.
for (g in c("extract_segments", "analyze_segment")) {
  methods <- grep(paste0("^", g, "[.]"), own, value = TRUE)
  defined[[g]] <- unique(c(setdiff(defined[[g]], "UseMethod"), methods))
}

render <- function(fn, prefix = "", seen = character(0), depth = 0L) {
  callees <- intersect(defined[[fn]], own)
  callees <- setdiff(callees, seen)
  if (depth >= 4L) callees <- character(0)
  out <- character(0)
  for (i in seq_along(callees)) {
    last <- i == length(callees)
    out <- c(out, paste0(prefix, if (last) "└─ " else "├─ ",
                         callees[[i]], "()"),
             render(callees[[i]], paste0(prefix, if (last) "   " else "│  "),
                    c(seen, callees), depth + 1L))
  }
  out
}

cat("audit_package()", render("audit_package", seen = "audit_package"),
    sep = "\n")
```

## Invariants

### Scanned files are never executed

Since pkgaudit is intended for reviewing untrusted packages, it must never
execute code contained in those packages. `tests/testthat/test-no_execution.R`
builds a package where every execution site writes a marker file, scans it,
exports it, and requires that none of the markers exist afterwards.

Three calls in the read path look like execution and are not.
`tools::parse_Rd()` parses without evaluating; `tools::loadPkgRdMacros()` reads
macro definitions as text; and `parse(text = )` produces an unevaluated
expression. A macro provider is looked up by macro name against a fixed list and
never taken from the audited package's `RdMacros` field, since the untrusted
package determines the field value.

### File size limits are applied before extraction

A file over 10 MB is not read. `extract_segments()` applies this restriction
before `UseMethod()` so that no method can skip it.
`tests/testthat/test-segments.R` verifies that an oversized file is not scanned
and that the refusal is recorded in `errors`.

### Failures are contained and counted

Listing files, reading a file, parsing code, and evaluating an XPath or a
regular expression are each wrapped in `tryCatch()`: a failure becomes a row in
`errors` rather than aborting the scan.

### Rules are integrity-checked

Rules live in a SQLite database built from YAML by `inst/scripts/build_db.R`.
`load_rules()` verifies the database against its SHA-256 sidecar on every call.
The check is time-of-check to time-of-use; see "Security considerations" in
`?load_rules`.

`load_rules()` also refuses a database with a gap in its phase model, so a rule
whose findings would carry no phases can never ship.

### Every rule is verified against its own examples

Each rule carries positive and negative examples in its YAML. `build_db.R`
writes them out as fixtures, and `tests/testthat/test-fixtures.R` requires every
positive to match its rule and every negative to not match.
