---
title: "Getting Started with pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit scans R source packages for files that can execute arbitrary commands
during autoconf, builds, checks, and installations, and for lifecycle hooks
whose bodies run automatically when a namespace is loaded, attached, unloaded,
or detached.
It also scans R source code for security-relevant patterns like `system()`
calls.

A finding is not an accusation. In many cases, flagged files and code will be
legitimate: files that can execute shell commands are needed for
system-dependent configuration, and many call system tools or download files for
good reason. pkgaudit helps you identify which parts of an untrusted package
should be reviewed before you install and load it.

```{r setup}
library(pkgaudit)
```

## pkgaudit Rules

pkgaudit separates *when* code executes from *what* code does using three rule
categories:

- **File contexts** are files that can execute arbitrary commands during
autoconf, builds, checks, and installations. These are primarily files
involved in compiling non-R source code, such as `configure`, `src/Makevars`,
and `src/Makefile`.
- **Code contexts** are lifecycle hooks whose bodies run automatically when a
namespace is loaded, attached, unloaded, or detached, such as `.onLoad()` and
`.onAttach()`.
- **Patterns** are security-relevant function calls. `system()`, for example,
can execute arbitrary shell commands. `source()` can fetch and execute a remote
payload.

Every pattern finding is attributed to the code context that contains it. A
`system()` call inside an ordinary function only runs if you call that function;
the same call inside `.onLoad()` runs automatically when a namespace is loaded.

The full rule set is documented in [pkgaudit Rule Coverage](rules.html).

## Rules and Database Integrity

The rules live in a versioned SQLite database shipped with the package.
`load_rules()` reads it, and `rules_version()` reports the version:

```{r rules}
rules_version()

rules <- load_rules()
nrow(rules$file_contexts)
nrow(rules$code_contexts)
nrow(rules$patterns)
nrow(rules$phases)
```

`phases` holds one row per context code can execute in: every file- and
code-context rule, plus `Top-level` and `Other`.

A modified database is one way to evade a scanner. `load_rules()` verifies the
database against its bundled SHA-256 sidecar on every call and refuses to load
a modified one. You can also check the hash of your installed copy against the
value published in the
[README](https://github.com/tylerjssmith/pkgaudit#database-integrity):

```{r hash}
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

## An Example Package

As a demonstration, we will scan `untrustedpkg`, a small source package shipped
with pkgaudit that does several things a reviewer would want to know about. It
is never built, checked, installed, or loaded; it exists only to be scanned.

```{r example-package}
tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

exdir <- file.path(tempdir(), "untrustedpkg-example")
utils::untar(tarball, exdir = exdir)

pkg <- file.path(exdir, "untrustedpkg")
list.files(pkg, recursive = TRUE)
```

Three of those files are worth reading before we scan them. A `configure` script
runs during installations from source, before any R source code is loaded:

```{r example-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

`.onLoad()` runs automatically when the namespace is loaded, so its body
executes on `library(untrustedpkg)`:

```{r example-onload, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "zzz.R")), sep = "\n")
```

An ordinary function only runs if the user calls it:

```{r example-fetch, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "fetch.R")), sep = "\n")
```

The script that generates this package is in
[data-raw/create_untrustedpkg.R](https://github.com/tylerjssmith/pkgaudit/blob/master/data-raw/create_untrustedpkg.R).

## Auditing a Package Directory

`audit_package()` scans an unpacked source package:

```{r audit}
result <- audit_package(pkg)
```

Printing the result gives the scan metadata and the number of findings by
category:

```{r print}
print(result)
```

`summary()` reports the findings themselves: how many run in each lifecycle
phase, which contexts were found, how often each pattern matched, and the MITRE
ATT&CK techniques involved.

```{r summary}
summary(result)
```

Both methods accept `path = FALSE`, which omits the local filesystem path from
the output. This is useful when sharing results, since the path may reveal a
username or directory layout.

A `pkgaudit` object is a named list of ordinary data frames, so you can filter,
join, and report on the findings however you like. The columns below are a
subset: each frame also carries the rule's `message`, too long to print legibly
and documented in [pkgaudit Rule Coverage](rules.html), the MITRE ATT&CK
techniques in `attck`, and one logical column per lifecycle phase.

```{r patterns-columns}
result$patterns[, c("rule", "file_context", "code_context", "line_number",
                    "column_number")]
```

Each finding also records when its code runs, as one logical column per phase:
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`on_load`, `on_attach`, `on_unload`, and `on_detach`. A pattern inside an
ordinary function is `FALSE` for every one of them -- it runs only if something
calls it -- and is counted in the `none` row of `summary()`.

```{r by-phase}
result$patterns[result$patterns$on_load,
                c("rule", "file_context", "code_context")]
```

## Auditing a Tarball

The more common workflow is to scan a package you have not yet installed.
`audit_tarball()` takes a `.tar.gz` source package, validates it, extracts it to
a temporary directory, and scans the result.

This is the `untrustedpkg` tarball we extracted above, scanned directly:

```{r tarball}
audit_tarball(tarball)
```

Because an untrusted archive is itself an attack surface, `audit_tarball()`
calls `validate_tar()` to check the tarball before its contents are extracted.
This check fails closed, refusing the whole archive, if it detects any link
entry, path traversal, absolute path, or malformed path or if the archive does
not extract to exactly one top-level directory. It also enforces limits on entry
count, uncompressed size, and compression ratio, guarding against archives that
expand to exhaust the disk.

```{r validate}
str(validate_tar(tarball))
```

## Interpreting a Scan

A clean scan is weak evidence of safety: pkgaudit reasons about syntax, so
sufficiently indirect code can evade any pattern rule. A scan with findings is
better understood as a reading list than a verdict. In review, the questions
worth asking are roughly:

- Does the package need this capability at all, given what it claims to do?
- Does it run without being asked, at install or load time, rather than when a
  user calls a function?
- Does it reach the network, and if so, is the destination fixed and
  identifiable in the source?
- Is the code being run visible in the source, or is it assembled, decoded, or
  fetched at runtime?

Findings that answer badly on several of these at once are the ones to read
closely.
