---
title: "Getting Started with pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit reports which parts of an R package can execute, when they execute,
and what they do, so that an untrusted package can be reviewed before it is
installed and loaded. It never executes the code it scans.

A finding is not an accusation. `configure` scripts and calls to system tools,
for example, are often legitimate. pkgaudit helps to identify what deserves
reviewer attention, not what is malicious.

```{r setup}
library(pkgaudit)
```

## Example package

`untrustedpkg` is a small source package shipped with pkgaudit. It is never
built, checked, installed, or loaded; it exists only to be scanned.

```{r example-package}
tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

exdir <- file.path(tempdir(), "untrustedpkg-example")
utils::untar(tarball, exdir = exdir)

pkg <- file.path(exdir, "untrustedpkg")
list.files(pkg, recursive = TRUE)
```

A `configure` script is a shell script that R runs when a package is installed
from source, before any of its R code is loaded:

```{r example-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

`R/zzz.R` defines `.onLoad()`, a hook R calls when the namespace is loaded:

```{r example-onload, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "zzz.R")), sep = "\n")
```

`R/fetch.R` defines an ordinary function:

```{r example-fetch, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "fetch.R")), sep = "\n")
```

A help file can carry R code in two places that run at different times. An
`\examples{}` block runs under `R CMD check`. A `\Sexpr{}` macro runs whenever
the page is rendered, which includes `R CMD build` and installation from
source:

```{r example-rd, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "man", "fetch_data.Rd")), sep = "\n")
```

The script that generates this package is in
[data-raw/create_untrustedpkg.R](https://github.com/tylerjssmith/pkgaudit/blob/main/data-raw/create_untrustedpkg.R).

## Auditing a package

`audit_package()` scans an unpacked source package. Printing the result gives
the scan metadata and the number of findings by category.

```{r audit}
result <- audit_package(pkg)

print(result)
```

`summary()` reports the findings themselves: how often each rule matched, split
by the lifecycle phase the code runs in, with the MITRE ATT&CK techniques the
rule carries.

```{r summary}
summary(result)
```

Both methods accept `path = FALSE`, which omits the local filesystem path. This
matters when sharing results, since the path may reveal a username or directory
layout.

A `pkgaudit` object is a named list of ordinary data frames, so findings can be
filtered, joined and reported on directly.

```{r names}
names(result)
```

`patterns` holds security-relevant R calls, located by file, line and column,
and records the code context each one sits in. Every findings frame also
carries the rule's `message`, its ATT&CK techniques, and one logical column per
lifecycle phase; those are omitted below to keep the output readable.

```{r patterns-columns}
result$patterns[, c("rule", "file_context", "line_number", "code_context",
                    "guarded", "indirect", "preview")]
```

`matches` mirrors `patterns` but carries no `code_context`: a shell script has
no R parse tree to sit in, so a match is located by file alone.

```{r matches-columns}
result$matches[, c("rule", "file_context", "line_number", "preview")]
```

`coverage` accounts for the files themselves, recording what pkgaudit made of
each one: `parsed` for R, matched against its parse tree; `matched` for shell
and Make-like files, matched as text; `exportable` for languages pkgaudit does
not read, such as C and Python; `unexamined` for files it did not read, such as
serialized `.rda` and `.rds` objects; and `error` where it tried to read a file
and could not.

```{r coverage}
result$coverage[, c("file_context", "language", "status", "reason", "lines")]
```

Coverage may not be complete. What the frame offers is not completeness but
legibility: a clean result can be checked rather than trusted, because the scan
says what it did not cover.

### Subsetting by phase

Every findings frame carries one logical column per lifecycle phase:
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`at_load`, `at_attach`, `at_unload`, `at_detach`. A pattern inside an ordinary
function is `FALSE` for all of them, since it runs only if something calls it.

```{r by-phase}
result$patterns[result$patterns$at_load,
                c("rule", "file_context", "code_context")]
```

`summary()` takes a `phase` argument for the same purpose, including `"none"`
for code that ships but runs at no phase.

```{r summary-phase}
summary(result, phase = "at_load", path = FALSE)
```

Phases overlap: building a package with vignettes also installs and loads it,
so one occurrence is counted under every phase it runs in.

## Reviewing findings

Two columns describe how code is *reached* rather than what it is, and both
bear on how much weight a finding deserves. `guarded` is `TRUE` for code that
ships but the lifecycle does not run, such as a `\dontrun{}` block or a
vignette chunk marked `eval=FALSE`. `indirect` is `TRUE` where a call was made
through the function's name rather than the function, as in
`do.call("system", ...)`; the finding is reported under the rule that owns the
name, so filtering on `rule` finds every call to it however it was spelled.

`untrustedpkg` returns five findings. Read in order of the phases they run in,
they escalate.

**`download.file()` in `R/fetch.R` runs at no phase.** It sits in the body of
`fetch_data()`, so it executes only if a user calls that function. A package
that downloads a file when asked to download a file is doing its job. This is
the baseline: a capability, disclosed in an exported function.

**`download.file()` in `man/fetch_data.Rd` runs under `R CMD check`.** The same
call appears in the `\examples{}` block, so checking the package fetches the
URL. Examples are meant to run, and a reviewer would weigh this as documentation
that reaches the network rather than as an attack.

**`system()` in `R/zzz.R` runs on `library()`.** `.onLoad()` is called when the
namespace loads, so `system("uname -a")` executes on attach, and again at build,
check, and installation from source, each of which loads the package. Nobody
asked for it. The command itself is reconnaissance rather than damage, but the
capability is arbitrary shell execution at load time.

**`curl` in `configure` runs at installation from source.** The script pipes a
remote script directly into a shell:

```{r review-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

This executes before any R code is loaded, and what it executes is not in the
package: it is whatever the remote host serves at the moment of installation.
Nothing in the source says what will run.

**`httr::POST()` in `man/fetch_data.Rd` runs when the help page is built, and
is invisible.** The call sits in a `\Sexpr[results=hide]{}` macro in the
`\description{}` block. `\Sexpr{}` is evaluated whenever the page is rendered,
during `R CMD build` and installation from source; `results=hide` suppresses
its output. A reader of `?fetch_data` sees a one-line description and an
example, and no sign that rendering the page sent `Sys.info()` to a remote
host.

```{r review-sexpr}
result$patterns[result$patterns$code_context == "Rd_Sexpr_install",
                c("rule", "file_context", "line_number", "preview")]
```

This is the finding that motivates scanning documentation as code. It runs
automatically, it exfiltrates, and it is invisible both to a user reading the
help page and to any scanner that reads only `.R` files.

A clean scan is weak evidence of safety. pkgaudit reasons about syntax, so
sufficiently indirect code can evade any pattern rule. A call whose target is a
string literal is followed -- `do.call("system", ...)` is reported as a `system`
finding -- but one assembled at runtime is not, since pkgaudit evaluates
nothing. Matches are weaker still, being matched against text rather than a
parse tree, so they both miss more and flag more: a `curl` inside a comment is
reported, and one assembled from shell variables is not.

A scan with findings is better understood as a reading list than a verdict. In
review, the questions worth asking are roughly:

- Does the package need this capability at all, given what it claims to do?
- Does it run without being asked, at install or load time, rather than when a
  user calls a function?
- Does it reach the network, and if so, is the destination fixed and
  identifiable in the source?
- Is the code being run visible in the source, or is it assembled, decoded, or
  fetched at runtime?
- What did the scan not read, and does any of it run automatically?

Findings that answer badly on several of these at once are the ones to read
closely.

## Exporting findings

`emit_sarif()` renders a result as SARIF 2.1.0, the format code-scanning tools
publish results in. It returns the document as a string and writes nothing.

```{r sarif}
sarif <- emit_sarif(result)
substr(sarif, 1, 200)
```

Written to a file and opened in an editor with a SARIF viewer, each finding
appears on the line it was found. Rule identifiers are namespaced by the kind of
rule -- `pattern/curl`, `match/curl` -- because a rule name is unique only
within its kind. `level` is `warning` for a pattern or match whose code executes
during at least one lifecycle phase and `note` for everything else. That is a
mapping of pkgaudit's phase model onto SARIF's severity field, not a severity
ranking, which pkgaudit does not make.

`export_unscanned()` writes the code pkgaudit cannot read into a directory a
scanner such as Semgrep can be pointed at. A whole file is copied verbatim; a
vignette chunk in a language with no analyser is written into a file of its own,
blank-padded so that its code sits at the same line numbers it occupies in the
source. A finding another tool reports at line 40 of `intro.python.py` is
therefore at line 40 of `intro.Rmd`.

`untrustedpkg` is R and shell throughout, so there is nothing to hand on and the
manifest comes back empty:

```{r export}
manifest <- export_unscanned(result, file.path(tempdir(), "for-semgrep"),
                             source = pkg)
nrow(manifest)
```

For a package carrying compiled code, each row of the manifest maps an exported
file back to where it came from. The directory has no default: naming one is how
the caller consents to being written to.

## Auditing a tarball

The more common workflow is to scan a package that has not been installed.
`audit_tarball()` takes a `.tar.gz` source package, validates it, extracts it to
a temporary directory, scans it, and removes the directory. An untrusted archive
is itself an attack surface, so validation fails closed: the whole archive is
refused rather than partially extracted.

```{r tarball}
print(audit_tarball(tarball), path = FALSE)
```

The result is a `pkgaudit` object like any other, so everything above applies to
it unchanged.

## Rule Database Integrity

The rules live in a versioned SQLite database shipped with the package.
`load_rules()` reads it; `rules_version()` reports the version.

```{r rules}
rules_version()

rules <- load_rules()
vapply(rules, nrow, integer(1))
```

A modified database is one way to evade a scanner. `load_rules()` verifies the
database against its bundled SHA-256 sidecar on every call and refuses to load
a modified one. The hash of an installed copy can also be checked against the
value published in the
[README](https://github.com/tylerjssmith/pkgaudit#database-integrity):

```{r hash}
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

The full rule set is documented in [Rule Coverage](rules.html). How pkgaudit
works internally is in [Internals](internals.html).
