---
title: "Getting Started with pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit scans R packages for security-relevant files and code without executing
anything it scans. It reports what code does and when it runs, so code that runs
on install or load is distinguishable from code that runs only when called.

A finding is not an accusation. Nearly all of the patterns and matches flagged
by pkgaudit have legitimate uses in R packages. pkgaudit helps to identify what
deserves reviewer attention, not what is malicious.

```{r setup}
library(pkgaudit)
```

## Example package

`untrustedpkg` is a small source package shipped with pkgaudit for
demonstration. It is never built, checked, installed, or loaded; it exists only
to be scanned.

```{r example-package}
tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

exdir <- file.path(tempdir(), "untrustedpkg-example")
utils::untar(tarball, exdir = exdir)
pkg <- file.path(exdir, "untrustedpkg")
```

`untrustedpkg` contains the following files.

```{r example-list, echo = FALSE, comment = ""}
list.files(pkg, recursive = TRUE)
```

The script that generates this package is in
[data-raw/create_untrustedpkg.R](https://github.com/tylerjssmith/pkgaudit/blob/main/data-raw/create_untrustedpkg.R).

## Auditing a package

`audit_package()` scans an untarred source package. `print()` gives the scan
metadata and the number of findings by category.

```{r audit}
result <- audit_package(pkg)
print(result, path = FALSE)
```

`summary()` reports the number of findings by rule by phase. It also
provides MITRE ATT&CK techniques associated with each rule.

```{r summary}
summary(result, path = FALSE)
```

Phases overlap -- building a package with vignettes, for example, also installs
and loads it -- and one occurrence is counted under every phase it runs in. The
summary above reflects five findings, some counted under multiple phases. To see
only what is known to run automatically at installation from source, for
example, pass an argument to `phase`:

```{r summary-phase}
summary(result, phase = c("at_install_src"), path = FALSE)
```

Both `print()` and `summary()` accept `path = FALSE`, which omits the local
filesystem path. This can matter when sharing results, since the path may reveal
a username or directory layout.

## Examining results

A `pkgaudit` object is a named list of ordinary data frames, plus scan metadata,
so findings can be filtered, joined and reported on directly.

```{r names}
names(result)
```

### File contexts

`$file_contexts` reports security-relevant files, whatever they contain: R runs
a `configure` script, which can execute shell commands, so the script should be
reviewed.

```{r file-contexts-columns}
result$file_contexts[, c("rule", "file_context")]
```

### Patterns

`$patterns` reports security-relevant R calls, located by `file_context`,
`line_number`, and `column_number`.

```{r patterns-columns1}
result$patterns[, c("rule", "file_context", "line_number", "column_number")]
```

Additionally, `code_context` indicates where code sits within a file. File and
code context are used to determine the phases. `guarded` is `TRUE` when a call
may be stopped from running in the expected phase, e.g., a `\dontrun{}` block or
a vignette marked `eval=FALSE`. `indirect` is `TRUE` when a call is made through
a function name, e.g., `do.call("system", ...)` instead of `system()`.

```{r patterns-columns2}
result$patterns[, c("rule", "code_context", "guarded", "indirect")]
```

`preview` provides a snippet of the code and its surroundings that may allow
reviewers to determine the relevance of a finding without opening the file.

```{r patterns-columns3}
result$patterns[, c("preview")]
```

### Matches

`$matches` reports text matching regular expressions in shell and Make-like
files. Its columns mirror `$patterns`, but it does not have `code_context`,
`guarded`, or `indirect`.

```{r matches-columns}
result$matches[, c("rule", "file_context", "line_number", "preview")]
```

### Coverage

`$coverage` reports what pkgaudit made of each file it scanned: `parsed` for R,
matched against its parse tree; `matched` for shell and Make-like files, matched
as text; `exportable` for languages pkgaudit does not read, such as C and Python;
`unexamined` for files it did not read, such as serialized `.rda` and `.rds`
objects; and `error` where it tried to read a file and could not.

```{r coverage}
result$coverage[, c("file_context", "language", "status", "reason", "lines")]
```

### Errors

`$errors` reports files that pkgaudit tried to read and could not, with
information about what may have gone wrong. pkgaudit did not encounter errors
when scanning `untrustedpkg`.

### Metadata

`$metadata` is a list of metadata.

```{r metadata}
result$metadata
```

### Subsetting by phase

Every findings frame carries one logical column per lifecycle phase:
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`at_load`, `at_attach`, `at_unload`, `at_detach`. A pattern inside an ordinary
function is `FALSE` for all of them, since it runs only if something calls it.
These can be used to subset the data frames and see what findings apply to each
phase. For example, to see what is known to run when `untrustedpkg` is loaded:

```{r by-phase}
result$patterns[result$patterns$at_load,
                c("rule", "file_context", "code_context")]
```

## Reviewing findings

pkgaudit is a guide to human review, not a substitute for human judgment. Below
we consider the four patterns and one match for `untrustedpkg`.

### Patterns

1. An `\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help
page is rendered at build, check, and source installation. Reviewers should
consider whether a help page should make an HTTP POST request and what it may
send to an external host.

2. The `\examples{}` block in the same help file calls `download.file()` at
check. Reviewers should verify what is downloaded. Since this depends on an
external host, and what is downloaded may change over time, reviewers should
also consider how the download is used -- for example, could another function
execute code if the file ever contained it?

```{r example-rd, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "man", "fetch_data.Rd")), sep = "\n")
```

3. `.onLoad()` in `R/zzz.R` calls `system()` and runs when users call
`library()`. It also runs at build, check, and source installation, each of
which loads the package. While `system()` calls are common in R packages, they
are less common in lifecycle hooks like `.onLoad()`, and reviewers should
inspect what commands would be run automatically on their systems.

```{r example-onload, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "zzz.R")), sep = "\n")
```

4. The `configure` script may invoke `curl` at build, check, and source
installation. Some R packages use `curl` to fetch dependencies that
cannot be vendored with the package. Reviewers should verify what is fetched and
what could happen if that changes.

```{r example-fetch, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "fetch.R")), sep = "\n")
```

### Matches

1. A regular function in `R/fetch.R` calls `download.file()`, but a regular
function is not known to run automatically, so this pattern is reported under
`none`. A reviewer may still want to inspect if and when the function is called,
and what would be downloaded.

```{r example-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

## Exporting findings

`emit_sarif()` renders a result as SARIF 2.1.0, the format code-scanning tools
publish results in. It returns the document as a string and writes nothing.

```{r sarif}
sarif <- emit_sarif(result)
substr(sarif, 1, 200)
```

Written to a file and opened in an editor with a SARIF viewer, each finding
appears on the line it was found. `level` is `note` for every result. When code
executes is carried in `properties.phases`.

`export_unscanned()` writes the code pkgaudit cannot read into a directory so
that a tool such as Semgrep can scan it. A whole file is copied verbatim; a
vignette chunk is written into a file of its own, blank-padded so that its code
sits at the same line numbers it occupies in the source. A finding another tool
reports at line 40 of `intro.python.py` is therefore at line 40 of `intro.Rmd`.
`untrustedpkg` contains only R and Bash, so `export_unscanned()` would create an
empty directory.

## Auditing a tarball

A common workflow will be to scan a package that has not been installed.
`audit_tarball()` takes a `.tar.gz` source package, validates it, extracts it to
a temporary directory, scans it, and removes the directory. An untrusted archive
is itself an attack surface, so validation fails closed: the whole archive is
refused rather than partially extracted.

```{r tarball}
print(audit_tarball(tarball), path = FALSE)
```

The result is a `pkgaudit` object like any other, so everything above applies to
it unchanged.

## Rule database integrity

The rules live in a versioned SQLite database shipped with the package.
`load_rules()` reads it; `rules_version()` reports the version.

```{r rules}
rules_version()

rules <- load_rules()
vapply(rules, nrow, integer(1))
```

A modified database is one way to evade a scanner. `load_rules()` verifies the
database against its bundled SHA-256 sidecar on every call and refuses to load
a modified one. The hash of an installed copy can also be checked against the
value published in the
[README](https://github.com/tylerjssmith/pkgaudit#rule-database-integrity):

```{r hash}
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

The full rule set is documented in [Rule Coverage](rules.html). How pkgaudit
works internally is in [Internals](internals.html).
