---
title: "Getting Started with pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit reports which parts of an R package can execute, when they execute,
and what they do, so that an untrusted package can be reviewed before it is
installed and loaded. It never executes the code it scans.

A finding is not an accusation. Flagged files and code are often legitimate:
`configure` scripts exist for system-dependent configuration, and many packages
call system tools or download files for good reason. pkgaudit identifies what
deserves a reader's attention, not what is malicious.

```{r setup}
library(pkgaudit)
```

## How a package is covered

pkgaudit accounts for every file it can identify as code and records what it
made of each one, in a `coverage` frame:

- **parsed** -- R, wherever a package carries it. Matched against R's parse
  tree, so a finding is located by line and column and attributed to the
  construct that encloses it.
- **matched** -- shell scripts and Make-like files, matched as text. Less
  precise: a match inside a comment reads the same as one in a live command.
- **exportable** -- C, C++, Fortran, Rust, Python, JavaScript, and vignette
  chunks in those languages. Not read by pkgaudit; `export_unscanned()` writes
  them out for a scanner that reads them.
- **unexamined** -- present and accounted for, but not read: serialized `.rda`
  and `.rds` objects, binaries, and files no rule claims.
- **error** -- pkgaudit tried to read the file and could not, because it was
  too large, unreadable, or would not parse. Distinct from `unexamined`, where
  it never tried.

Coverage is never complete, and is not meant to be. What the frame offers is
not completeness but legibility: a clean result can be checked rather than
taken on trust, because the scan says what it did not look at.

## Rules

pkgaudit separates *when* code executes from *what* code does, using four
categories of rule.

- **File contexts** tell the scan which files to read, and how. Most exist so
  that a file's *contents* can be reported -- `R/`, `man/`, `vignettes/`,
  `tests/`, `data/`. A file context is a finding in its own right only when it
  both runs automatically *and* can only be matched as text: in practice the
  shell scripts and Make-like files, where pkgaudit cannot say what the file
  does, only that it runs, so the file itself is what needs reading.
- **Code contexts** say when a piece of R code runs. Some are lifecycle hooks
  matched by a rule -- `.onLoad()`, `.onAttach()` and their kin. The rest are
  computed from where the code sits: the top level of `R/`, a help-file
  example, a `\Sexpr{}` macro at a given stage, a vignette chunk, a test.
- **Patterns** are security-relevant function calls, matched against R's parse
  tree. `system()` executes shell commands; `source()` can fetch and execute a
  remote script.
- **Matches** are regular-expression matches in shell scripts and Make-like
  files. `curl` and `wget`, for example, can fetch a remote payload or send
  data to a remote host while a package is being built or installed.

A code context is not a finding of its own. It travels on `patterns`, as the
column a finding's lifecycle phases are derived from.

The full rule set is documented in [Rule Coverage](rules.html).

## Rules and database integrity

The rules live in a versioned SQLite database shipped with the package.
`load_rules()` reads it; `rules_version()` reports the version.

```{r rules}
rules_version()

rules <- load_rules()
vapply(rules, nrow, integer(1))
```

`phases` holds one row per context code can execute in: every file- and
code-context rule, plus the computed contexts.

Not every file-context rule is a finding. Rules covering `R/` and `man/` tell
the scan which files to read; only a rule with `report = TRUE` contributes a
row to `result$file_contexts`.

```{r reporting-rules}
table(rules$file_contexts$type, rules$file_contexts$report)
```

A modified database is one way to evade a scanner. `load_rules()` verifies the
database against its bundled SHA-256 sidecar on every call and refuses to load
a modified one. The hash of an installed copy can also be checked against the
value published in the
[README](https://github.com/tylerjssmith/pkgaudit#database-integrity):

```{r hash}
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

## An example package

`untrustedpkg` is a small source package shipped with pkgaudit that does three
things a reviewer would want to know about. It is never built, checked,
installed, or loaded; it exists only to be scanned.

```{r example-package}
tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

exdir <- file.path(tempdir(), "untrustedpkg-example")
utils::untar(tarball, exdir = exdir)

pkg <- file.path(exdir, "untrustedpkg")
list.files(pkg, recursive = TRUE)
```

A `configure` script runs during installation from source, before any R code is
loaded. This one fetches a remote script and pipes it into a shell:

```{r example-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

`.onLoad()` runs automatically when the namespace is loaded, so its body
executes on `library(untrustedpkg)`. An ordinary function, by contrast, runs
only if someone calls it:

```{r example-onload, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "zzz.R")), sep = "\n")
cat("\n")
cat(readLines(file.path(pkg, "R", "fetch.R")), sep = "\n")
```

A help file carries R code too, in two places that run at different times. Its
`\examples{}` block runs under `R CMD check`. Its `\Sexpr{}` macro runs whenever
the page is rendered -- during `R CMD build` and installation from source -- and
here it is written `\Sexpr[results=hide]{...}`, which suppresses the output. A
person reading `?fetch_data` sees the description and the example, but nothing
of the `httr::POST()` call: it runs on their machine, invisibly, when the page
is built.

```{r example-rd, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "man", "fetch_data.Rd")), sep = "\n")
```

The script that generates this package is in
[data-raw/create_untrustedpkg.R](https://github.com/tylerjssmith/pkgaudit/blob/master/data-raw/create_untrustedpkg.R).

## Scanning a package directory

`audit_package()` scans an unpacked source package. Printing the result gives
the scan metadata and the number of findings by category.

```{r audit}
result <- audit_package(pkg)

print(result)
```

`summary()` reports the findings themselves: how often each rule matched, split
by the lifecycle phase the code runs in, with the MITRE ATT&CK techniques the
rule carries. Phases overlap -- building a package with vignettes also installs
and loads it -- so one occurrence is counted under every phase it runs in.

```{r summary}
summary(result)
```

Both methods accept `path = FALSE`, which omits the local filesystem path. This
matters when sharing results, since the path may reveal a username or directory
layout.

## Reading the result

A `pkgaudit` object is a named list of ordinary data frames, so findings can be
filtered, joined and reported on directly.

```{r names}
names(result)
```

`patterns` locates each finding by file, line and column, and records the code
context it sits in. Each frame also carries the rule's `message`, its ATT&CK
techniques, and one logical column per lifecycle phase; those are omitted below
to keep the output readable.

```{r patterns-columns}
result$patterns[, c("rule", "file_context", "line_number", "code_context",
                    "guarded", "indirect", "preview")]
```

Two columns describe how the code is *reached* rather than what it is.
`guarded` is `TRUE` for code that ships but the lifecycle does not run -- a
`\dontrun{}` block, or a vignette chunk marked `eval=FALSE`. `indirect` is
`TRUE` where the call was made through the function's name rather than the
function, as in `do.call("system", ...)`; such a finding is reported under the
rule that owns the name, so filtering on `rule` finds every call to it however
it was spelled.

`matches` mirrors `patterns` but carries no `code_context`: a shell script has
no R parse tree to sit in, so a match is located by file alone.

```{r matches-columns}
result$matches[, c("rule", "file_context", "line_number", "preview")]
```

`coverage` accounts for the files themselves.

```{r coverage}
result$coverage[, c("file_context", "language", "status", "reason", "lines")]
```

### Filtering by phase

Every findings frame carries one logical column per phase: `at_autoconf`,
`at_build`, `at_check`, `at_install_src`, `at_install_bin`, `at_load`,
`at_attach`, `at_unload`, `at_detach`. A pattern inside an ordinary function is
`FALSE` for all of them -- it runs only if something calls it.

```{r by-phase}
result$patterns[result$patterns$at_load,
                c("rule", "file_context", "code_context")]
```

`summary()` takes a `phase` argument for the same purpose, including `"none"`
for code that ships but runs at no phase.

```{r summary-phase}
summary(result, phase = "at_load", path = FALSE)
```

## Scanning a tarball

The more common workflow is to scan a package that has not been installed.
`audit_tarball()` validates a `.tar.gz` source package, extracts it to a
temporary directory, scans it, and removes the directory.

```{r tarball}
print(audit_tarball(tarball), path = FALSE)
```

Because an untrusted archive is itself an attack surface, the tarball is
checked before anything is extracted. The check fails closed, refusing the whole
archive, on any link entry, path traversal, absolute or malformed path, or an
archive that does not extract to exactly one top-level directory. It also
enforces limits on entry count, uncompressed size, and compression ratio,
guarding against archives that expand to exhaust the disk.

```{r validate}
str(validate_tar(tarball))
```

## Carrying a scan to other tools

`emit_sarif()` renders a result as SARIF 2.1.0, the format code-scanning tools
publish results in. It returns the document as a string and writes nothing.

```{r sarif}
sarif <- emit_sarif(result)
substr(sarif, 1, 200)
```

Written to a file and opened in an editor with a SARIF viewer, each finding
appears on the line it was found. Rule identifiers are namespaced by the kind of
rule -- `pattern/curl`, `match/curl` -- because a rule name is unique only
within its kind. `level` is `warning` for a pattern or match whose code executes
during at least one lifecycle phase and `note` for everything else. That is a
mapping of pkgaudit's phase model onto SARIF's severity field, not a severity
ranking, which pkgaudit does not make.

`export_unscanned()` writes the code pkgaudit cannot read into a directory a
scanner such as Semgrep can be pointed at. A whole file is copied verbatim; a
vignette chunk in a language with no analyser is written into a file of its own,
blank-padded so that its code sits at the same line numbers it occupies in the
source. A finding another tool reports at line 40 of `intro.python.py` is
therefore at line 40 of `intro.Rmd`.

`untrustedpkg` is R and shell throughout, so there is nothing to hand on and the
manifest comes back empty:

```{r export}
manifest <- export_unscanned(result, file.path(tempdir(), "for-semgrep"),
                             source = pkg)
nrow(manifest)
```

For a package carrying compiled code, each row of the manifest maps an exported
file back to where it came from. The directory has no default: naming one is how
the caller consents to being written to.

## Interpreting a scan

A clean scan is weak evidence of safety. pkgaudit reasons about syntax, so
sufficiently indirect code can evade any pattern rule. A call whose target is a
string literal is followed -- `do.call("system", ...)` is reported as a `system`
finding -- but one assembled at runtime is not, since pkgaudit evaluates
nothing. Matches are weaker still, being matched against text rather than a
parse tree, so they both miss more and flag more: a `curl` inside a comment is
reported, and one assembled from shell variables is not.

A scan with findings is better understood as a reading list than a verdict. In
review, the questions worth asking are roughly:

- Does the package need this capability at all, given what it claims to do?
- Does it run without being asked, at install or load time, rather than when a
  user calls a function?
- Does it reach the network, and if so, is the destination fixed and
  identifiable in the source?
- Is the code being run visible in the source, or is it assembled, decoded, or
  fetched at runtime?
- What did the scan not read, and does any of it run automatically?

Findings that answer badly on several of these at once are the ones to read
closely.
