---
title: "Getting Started with pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit is a static analysis security testing (SAST) tool for R source
packages. It flags security-relevant files and R source code for manual review, 
and organizes what it finds by the point in the R package lifecycle at which the
code runs (e.g., build, check, install, load).

A finding is not an accusation. Most flagged code is legitimate: packages
compile source code, download data, and call shell tools for good reasons. The
purpose of a scan is to help you identify *which* parts of an unfamiliar package
deserve your attention before you install and load it.

```{r setup}
library(pkgaudit)
```

## pkgaudit Rules

pkgaudit v0.3.0 separates *when* code executes from *what* code does using three
categories of rules:

- **File contexts** are files that R executes at build-, check-, or 
install-time, such as `configure` and `src/Makevars`.
- **Code contexts** are lifecycle hooks whose bodies run automatically when a
namespace is loaded, attached, unloaded, or detached, such as `.onLoad()` and
`.onAttach()`.
- **Patterns** are security-relevant function calls, such as `system()`, which
can execute arbitrary shell commands.

The three combine to give a finding its weight. A `system()` call inside an
ordinary function only runs if you call that function; the same call inside
`.onLoad()` runs the moment the namespace loads. Every pattern finding is
therefore attributed to the code context that contains it.

The full rule set is documented in
[RULES.md](https://github.com/tylerjssmith/pkgaudit/blob/master/RULES.md).

## An Example Package

As a demonstration, we will scan `untrustedpkg`, a small source package shipped
with pkgaudit that does several things a reviewer would want to know about. It
is never built, installed, or loaded; it exists only to be scanned.

```{r example-package}
tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

exdir <- file.path(tempdir(), "untrustedpkg-example")
utils::untar(tarball, exdir = exdir)

pkg <- file.path(exdir, "untrustedpkg")
list.files(pkg, recursive = TRUE)
```

Three of those files are worth reading before we scan them. A `configure` script
runs at install time, before any R code is loaded:

```{r example-configure, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "configure")), sep = "\n")
```

`.onLoad()` runs automatically when the namespace is loaded, so its body
executes on `library(untrustedpkg)`:

```{r example-onload, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "zzz.R")), sep = "\n")
```

An ordinary function only runs if the user calls it:

```{r example-fetch, echo = FALSE, comment = ""}
cat(readLines(file.path(pkg, "R", "fetch.R")), sep = "\n")
```

The script that generates this package is in
[data-raw/create_untrustedpkg.R](https://github.com/tylerjssmith/pkgaudit/blob/master/data-raw/create_untrustedpkg.R).

## Auditing a Package Directory

`audit_package()` scans an unpacked source package:

```{r audit}
result <- audit_package(pkg)
```

Printing the result gives the scan metadata and the number of findings by
category:

```{r print}
result
```

`summary()` reports the findings themselves: which contexts were found, how
often each pattern matched, and the MITRE ATT&CK techniques involved.

```{r summary}
summary(result)
```

Both methods accept `path = FALSE`, which omits the local filesystem path from
the output. This is useful when sharing results, since the path may reveal a
username or directory layout.

```{r print-nopath}
print(result, path = FALSE)
```

## Working with pkgaudit Results

A `pkgaudit` object is a named list of ordinary data frames, so you can filter,
join, and report on the findings however you like.

Every findings frame also carries one logical column per lifecycle phase, which
the next section covers; the columns shown here are the rest.

```{r components}
result$file_contexts[, 1:3]
result$code_contexts[, 1:5]
result$patterns[, 1:7]
```

Each row of `result$patterns` carries the code context it was found in, so the
distinction that matters most is a single subset away:

```{r at-load}
result$patterns[result$patterns$code_context == "onload_code", 1:7]
```

## Filtering by Lifecycle Phase

Each finding also records when its code runs, as one logical column per phase:
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`on_load`, `on_attach`, `on_unload`, and `on_detach`. A file or code context
takes these from the rule that matched it; a pattern inherits them from its code
context.

That makes "what runs if I merely install this package?" a subset:

```{r by-phase}
result$patterns[result$patterns$at_install_src,
                c("rule", "file_context", "code_context")]
```

A finding can run in several phases, so the columns do not partition the rows. A
pattern inside an ordinary function is `FALSE` for every phase: it runs only if
something calls it, which is a materially weaker claim on your attention than
code that runs on installation. The `Findings by Phase` section of `summary()`
counts findings by phase, with a trailing `none` row for exactly that case.

Two further components round out the object. `result$errors` records any file or
rule that could not be processed, so a parse failure is reported rather than
silently dropped; an empty data frame means the scan was complete.
`result$metadata` records what was scanned and with what.

```{r metadata}
result$errors
str(result$metadata)
```

## Auditing a Tarball

The more common workflow is to scan a package you have not yet installed.
`audit_tarball()` takes a `.tar.gz` source package, validates it, extracts it to
a temporary directory, and scans the result.

This is the `untrustedpkg` tarball we extracted above, scanned directly:

```{r tarball}
audit_tarball(tarball)
```

Because an untrusted archive is itself an attack surface, it is checked before
anything is written to disk. `validate_tar()` performs that check on its own. It
fails closed, refusing the whole archive rather than extracting a filtered
subset: any link entry, path traversal, absolute path, or malformed path is
grounds for refusal, as is an archive that does not extract to exactly one
top-level directory. It also enforces limits on entry count, uncompressed size,
and compression ratio, guarding against archives that expand to exhaust the
disk:

```{r validate}
str(validate_tar(tarball))
```

## Rules and Database Integrity

The rules live in a versioned SQLite database shipped with the package.
`load_rules()` reads it, and `rules_version()` reports the version:

```{r rules}
rules_version()

rules <- load_rules()
nrow(rules$file_contexts)
nrow(rules$code_contexts)
nrow(rules$patterns)
nrow(rules$phases)
```

`phases` holds one row per context code can execute in: every file- and
code-context rule, plus `Top-level` and `Other`.

Passing the rules explicitly avoids re-reading and re-verifying the database on
every call, which is worthwhile when scanning many packages:

```{r reuse-rules, eval = FALSE}
rules   <- load_rules()
results <- lapply(tarballs, audit_tarball, rules = rules)
```

Because the rules determine what a scan can find, a modified database is a way
to make a scan come back clean. `load_rules()` verifies the database against its
bundled SHA-256 sidecar on every call and refuses to load a modified one. You
can also check the hash of your installed copy against the value published in
the [README](https://github.com/tylerjssmith/pkgaudit#database-integrity):

```{r hash}
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

## Interpreting a Scan

A clean scan is weak evidence of safety: pkgaudit reasons about syntax, so
sufficiently indirect code can evade any pattern rule. A scan with findings is
better understood as a reading list than a verdict. In review, the questions
worth asking are roughly:

- Does the package need this capability at all, given what it claims to do?
- Does it run without being asked, at install or load time, rather than when a
  user calls a function?
- Does it reach the network, and if so, is the destination fixed and
  identifiable in the source?
- Is the code being run visible in the source, or is it assembled, decoded, or
  fetched at runtime?

Findings that answer badly on several of these at once are the ones to read
closely.
