---
title: "pkgaudit Rule Coverage"
output:
  rmarkdown::html_vignette:
    # Rule messages quote shell flags and make constructs verbatim, so the
    # "smart" typography that would turn --clean into an en dash, or
    # $(shell ...) into an ellipsis, is disabled.
    md_extensions: "-smart"
vignette: >
  %\VignetteIndexEntry{pkgaudit Rule Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  results = "asis"
)
library(pkgaudit)

# Read the installed database so the tables below cannot drift from the rules
# the package actually ships.
rules       <- load_rules()
phase_names <- setdiff(names(rules$phases), c("context", "version"))
```

```{r helpers}
# The rule sources are read from the installed package, but linked in the
# repository, where a reader can browse them.
repo <- "https://github.com/tylerjssmith/pkgaudit/blob/main/"

# Map each rule name to the YAML file that defines it, as a path relative to
# inst/rules/. The database records a rule's name but not where it came from,
# so the rule sources are read for it.
#
# Keyed by category as well as name: a rule name is unique only within its
# category, and `curl` names both a pattern rule and a match rule.
rule_files <- local({
  root  <- system.file("rules", package = "pkgaudit")
  paths <- list.files(root, pattern = "\\.yaml$", recursive = TRUE)
  named <- vapply(paths, function(path) {
    line <- grep("^name: ", readLines(file.path(root, path)), value = TRUE)
    if (length(line) == 0L) NA_character_ else sub("^name: ", "", line[[1L]])
  }, character(1L))
  # The computed contexts have no rule file to link to, and
  # carry no name field.
  keep <- !is.na(named)
  split(stats::setNames(paths[keep], named[keep]), dirname(paths[keep]))
})

# A rule name, linked to the YAML that defines it in the given category.
rule_link <- function(name, category) {
  sprintf("[%s](%sinst/rules/%s)", name, repo, rule_files[[category]][name])
}

# The phases a context runs in, named as the columns a caller would filter on.
# "none" is what `summary()` calls a context that runs at no phase at all.
phase_cell <- function(context) {
  row <- rules$phases[match(context, rules$phases$context), phase_names]
  on  <- phase_names[as.logical(row)]
  if (length(on) == 0L) "none" else paste0("`", on, "`", collapse = ", ")
}

# A file context's file name, recovered from the rule's search path and
# pattern so the table cannot disagree with what the rule matches.
#
# A rule anchored at both ends names one file. A rule matching a class of files
# -- every R source in R/, say -- has no single name to show, so its pattern is
# shown after a wildcard instead.
anchored <- function(pattern) grepl("^\\^", pattern) & grepl("\\$$", pattern)

file_label <- function(path, pattern) {
  literal <- gsub("\\\\", "", sub("\\$$", "", sub("^\\^", "", pattern)))
  file    <- ifelse(anchored(pattern), literal, paste0("*", literal))
  ifelse(path == ".", file, paste0(path, "/", file))
}

# Render a markdown table, one row per line, escaping any cell content that
# would otherwise be read as a column separator.
emit_table <- function(headers, ...) {
  cells <- lapply(list(...), function(x) gsub("|", "\\|", x, fixed = TRUE))
  cat(
    c(
      paste0("| ", paste(headers, collapse = " | "), " |"),
      paste0("|", strrep("---|", length(headers))),
      paste0("| ", do.call(paste, c(cells, sep = " | ")), " |")
    ),
    sep = "\n"
  )
  cat("\n\n")
}
```

```{r labels}
# What each code-context and pattern rule covers, in the terms a reader would
# recognize. These cannot be derived from the rule: an XPath expression is a
# definition, not a summary of one. Every rule in the database must appear
# here, so a new rule fails this document until it is described.
code_labels <- c(
  LastLib_base     = "`.Last.lib()`",
  onAttach_base    = "`.onAttach()`",
  onDetach_base    = "`.onDetach()`",
  onLoad_base      = "`.onLoad()`",
  on_load_rlang    = "`rlang::on_load()`",
  onUnload_base    = "`.onUnload()`",
  Rd_examples      = "`\\examples{}`",
  Rd_Sexpr_build   = "`\\Sexpr[stage=build]`",
  Rd_Sexpr_install = "`\\Sexpr[stage=install]`",
  Rd_Sexpr_render  = "`\\Sexpr[stage=render]`"
)

pattern_labels <- c(
  chmod = "`Sys.chmod()`, `Sys.umask()`",
  credentials = paste(
    "a path or environment variable that only holds a secret: `~/.ssh/`,",
    "`~/.aws/`, `.netrc`, `id_rsa`, `~/.gnupg/`, and `Sys.getenv()` of a name",
    "containing `TOKEN`, `SECRET`, `PASSWORD` or `API_KEY`"
  ),
  curl = paste(
    "`curl()`, `curl_fetch_memory()`, `curl_fetch_disk()`,",
    "`curl_fetch_stream()`, `curl_fetch_multi()`, `curl_download()`,",
    "`curl_upload()`, `multi_download()`, `multi_run()`, `send_mail()`"
  ),
  decoding = paste(
    "`base64decode()`, `base64_dec()`, `base64_decode()`, `memDecompress()`,",
    "`rawToChar(as.raw())`"
  ),
  deserialization =
    "`readRDS()`, `load()`, `unserialize()`, `dget()`",
  download_file = "`download.file()`, `url()`",
  dynload       = "`dyn.load()`, `library.dynam()`",
  eval_parse = paste(
    "`eval()`/`evalq()` over `parse()`, `str2lang()` or `str2expression()`,",
    "when the text came from a fetch, a decode, a deserialization, a",
    "subprocess or a file read -- most functions the rules for those already",
    "cover, plus `readLines()`, `scan()`, `readBin()` and `rawToChar()`;",
    "callr's and processx's generically named entry points are not matched"
  ),
  fs   = "`file_chmod()`, `file_chown()`",
  httr = paste(
    "`GET()`, `POST()`, `PUT()`, `PATCH()`, `DELETE()`, `HEAD()`, `VERB()`"
  ),
  httr2 = paste(
    "`req_perform()`, `req_perform_iterative()`, `req_perform_parallel()`,",
    "`req_perform_sequential()`, `req_perform_stream()`,",
    "`req_perform_connection()`, `req_perform_promise()`"
  ),
  install = paste(
    "`install.packages()`, `install_github()`, `install_gitlab()`,",
    "`install_bitbucket()`, `install_git()`, `install_url()`,",
    "`install_version()`, `install_local()`, `install_bioc()`,",
    "`install_dev()`, `install_cran()`, `pak()`, `pkg_install()`,",
    "`local_install()`, `lockfile_install()`, `BiocManager::install()`,",
    "`renv::install()`, `devtools::install()`, `py_install()`,",
    "`virtualenv_install()`, `conda_install()`, `virtualenv_create()`,",
    "`conda_create()`"
  ),
  namespace = paste(
    "`getFromNamespace()`, `getExportedValue()`, `getAnywhere()`,",
    "`assignInNamespace()`, `assignInMyNamespace()`, `unlockBinding()`"
  ),
  options_repos = "`options(repos = )`",
  persistence = paste(
    "a startup file or scheduler: `~/.bashrc`, `~/.zshrc`, `~/.profile`,",
    "`~/.Rprofile`, `~/.Renviron`, `crontab`, `LaunchAgents`, `/etc/systemd`"
  ),
  python = paste(
    "`py_run_string()`, `py_run_file()`, `py_eval()`, `source_python()`,",
    "`py_call()`, `import_builtins()`"
  ),
  rcurl = paste(
    "`getURL()`, `getURI()`, `getForm()`, `postForm()`, `curlPerform()`"
  ),
  socket = paste(
    "`socketConnection()`, `make.socket()`, `serverSocket()`,",
    "`socketAccept()`"
  ),
  source = "`source()`",
  system = "`system()`, `system2()`, `shell()`, `pipe()`",
  system_callr = paste(
    "`callr::r()`, `r_bg()`, `r_safe()`, `r_copycat()`, `r_vanilla()`,",
    "`callr::rcmd()`, `rcmd_bg()`, `rcmd_safe()`, `rcmd_copycat()`,",
    "`rscript()`, `r_session$new()`, `r_process$new()`,",
    "`rcmd_process$new()`, `rscript_process$new()`"
  ),
  system_processx = paste(
    "`processx::run()`, `processx::pipeline()`, `process$new()`"
  ),
  system_sys = paste(
    "`exec_wait()`, `exec_background()`, `exec_internal()`, `eval_safe()`,",
    "`eval_fork()`, `r_background()`, `r_internal()`, `r_wait()`"
  )
)

# The two computed code contexts, which no rule defines and which carry no
# phases of their own.
computed_contexts <- data.frame(
  context = c("top_level", "in_function"),
  description = c(
    paste(
      "Code outside any function definition. It carries the phases of the file",
      "context it sits in, so the same call reads at_check under tests/ and",
      "at_build under data/."
    ),
    paste(
      "Code inside a function definition. It inherits the phases of the code",
      "around it, except under R/, where the rules report it as running at no",
      "phase."
    )
  ),
  stringsAsFactors = FALSE
)

# Every rule and computed context must be described above, and every file
# context's derived file name must be matched by the pattern it came from.
stopifnot(
  all(rules$file_contexts$name %in% names(rule_files$file_contexts)),
  all(rules$code_contexts$name %in% names(rule_files$code_contexts)),
  all(rules$patterns$name      %in% names(rule_files$patterns)),
  all(rules$matches$name       %in% names(rule_files$matches)),
  setequal(names(code_labels), rules$code_contexts$name),
  setequal(names(pattern_labels), rules$patterns$name),
  # The computed contexts are exactly those with no phases row: they inherit.
  !any(computed_contexts$context %in% rules$phases$context),
  length(setdiff(rules$phases$context,
                 c(rules$file_contexts$name, rules$code_contexts$name))) == 0L,
  # Every code context a file-context rule names must be a rule that exists.
  all(setdiff(unlist(strsplit(
        stats::na.omit(rules$file_contexts$code_context), "[[:space:]]+")),
      "computed") %in% rules$code_contexts$name),
  # Only a rule naming one file has a name to check against its own pattern.
  local({
    fc  <- rules$file_contexts[anchored(rules$file_contexts$filename), ]
    all(mapply(grepl, fc$filename, basename(file_label(fc$path, fc$filename))))
  })
)
```

In pkgaudit v`r utils::packageVersion("pkgaudit")`, pattern and match rules
describe what code does; phases resolved from file-context and code-context
rules describe when it runs. All rules are defined in YAML files under
[inst/rules/](https://github.com/tylerjssmith/pkgaudit/tree/main/inst/rules) and
compiled into the SQLite database at `inst/db/rules.db`. This vignette is
generated from that database, at rules v`r rules_version()`; the Rule columns
below link to the defining YAML.

## Patterns

Patterns are security-relevant function calls such as `system()` and strings
such as "`~/.ssh`". Qualified (`pkg::fn()`, `pkg:::fn()`) and unqualified
(`fn()`) call forms are both detected. A rule that declares its function names
also flags calls made through a name -- `do.call("system", ...)`,
`match.fun("system")`, and `getFunction("system")`. Pattern rules carry
[MITRE ATT&CK](https://attack.mitre.org/) technique labels.

```{r patterns}
pat <- rules$patterns
emit_table(
  c("Rule", "Pattern", "Description"),
  rule_link(pat$name, "patterns"),
  pattern_labels[pat$name],
  pat$message
)
```

## Matches

Matches are regular-expression matches in shell code: the shell scripts and
Make-like files among the [file contexts](#file-contexts) below, and the
`bash`, `sh`, and `zsh` chunks a vignette yields. Regular expressions are
matched with `gregexpr(perl = TRUE)` and are case-sensitive. Match rules carry
[MITRE ATT&CK](https://attack.mitre.org/) technique labels.

Note: Matching text is less precise than matching a parse tree. A match has no
syntax behind it, so a match inside a comment, a quoted string, or a branch that
never runs is reported the same as one in a live command.

```{r matches}
rx <- rules$matches
emit_table(
  c("Rule", "Description"),
  rule_link(rx$name, "matches"),
  rx$message
)
```

## Phases and Contexts

### Phases

Phases describe when code executes. For pattern findings, they are resolved from
file and code contexts. For match findings, they are resolved from file
contexts alone. See [Internals](internals.html).

```{r phases-legend}
phase_descriptions <- c(
  at_autoconf    = "Autoconf is run to generate `configure`",
  at_build       = "`R CMD build`",
  at_check       = "`R CMD check`",
  at_install_src = "`R CMD INSTALL` from source",
  at_install_bin = "a prebuilt binary package is installed",
  at_load        = "the namespace is loaded",
  at_attach      = "the package is attached to the search path",
  at_unload      = "the namespace is unloaded",
  at_detach      = "the package is detached from the search path"
)
stopifnot(setequal(names(phase_descriptions), phase_names))

emit_table(
  c("Phase", "Code executes when"),
  paste0("`", phase_names, "`"),
  phase_descriptions[phase_names]
)
```

### File Contexts

File contexts are files that may contain code. All are scanned, and some
security-relevant files are reported as findings in their own right. In general,
shell scripts and Make-like files as well as `src/install.libs.R` are reported.

```{r file-contexts}
fc <- rules$file_contexts
emit_table(
  c("Rule", "Report", "Phases", "Description"),
  rule_link(fc$name, "file_contexts"),
  ifelse(fc$report, "yes", "no"),
  vapply(fc$name, phase_cell, character(1L)),
  fc$message
)
```

### Code Contexts

Code contexts indicate where code sits within its file context. It is either
defined by a rule or computed. For rule-based code contexts, a lifecycle hook
like `.onLoad()` is found by matching the XML parse tree, while part of a help
file like `\examples` or `\Sexpr` is found by matching a label from the
extraction method. See [Internals](internals.html). If no code-context rule
matches, code context is computed from the parse tree: `in_function` when code
sits inside a regular function definition, and `top_level` otherwise.

#### Rule-based

```{r code-contexts-rules}
cc <- rules$code_contexts
emit_table(
  c("Rule", "Code Context", "Matched by", "Phases", "Description"),
  rule_link(cc$name, "code_contexts"),
  code_labels[cc$name],
  ifelse(cc$kind == "xpath", "parse tree", "segment label"),
  vapply(cc$name, phase_cell, character(1L)),
  cc$message
)
```

#### Computed

```{r code-contexts-computed}
emit_table(
  c("Context", "Phases", "Description"),
  paste0("`", computed_contexts$context, "`"),
  rep("inherited from file context", nrow(computed_contexts)),
  computed_contexts$description
)
```
