---
title: "pkgaudit Rule Coverage"
output:
  rmarkdown::html_vignette:
    # Rule messages quote shell flags and make constructs verbatim, so the
    # "smart" typography that would turn --clean into an en dash, or
    # $(shell ...) into an ellipsis, is disabled.
    md_extensions: "-smart"
vignette: >
  %\VignetteIndexEntry{pkgaudit Rule Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "asis")

library(pkgaudit)

# Read the installed database so the tables below cannot drift from the rules
# the package actually ships.
rules       <- load_rules()
phase_names <- setdiff(names(rules$phases), c("context", "version"))
```

```{r helpers}
# The rule sources are read from the installed package, but linked in the
# repository, where a reader can browse them.
repo <- "https://github.com/tylerjssmith/pkgaudit/blob/master/"

# Map each rule name to the YAML file that defines it, as a path relative to
# inst/rules/. The database records a rule's name but not where it came from,
# so the rule sources are read for it.
rule_files <- local({
  root  <- system.file("rules", package = "pkgaudit")
  paths <- list.files(root, pattern = "\\.yaml$", recursive = TRUE)
  named <- vapply(paths, function(path) {
    line <- grep("^name: ", readLines(file.path(root, path)), value = TRUE)
    if (length(line) == 0L) NA_character_ else sub("^name: ", "", line[[1L]])
  }, character(1L))
  # Files under inst/rules/phases/ define a computed context, not a rule, and
  # carry no name field.
  stats::setNames(paths[!is.na(named)], named[!is.na(named)])
})

# A rule name, linked to its YAML.
rule_link <- function(name) {
  sprintf("[%s](%sinst/rules/%s)", name, repo, rule_files[name])
}

# The phases a context runs in, named as the columns a caller would filter on.
# "none" is the reading `summary()` gives it: code reached only when something
# calls it.
phase_cell <- function(context) {
  row <- rules$phases[match(context, rules$phases$context), phase_names]
  on  <- phase_names[as.logical(row)]
  if (length(on) == 0L) "none" else paste0("`", on, "`", collapse = ", ")
}

# A file context's file name, recovered from the rule's search path and
# anchored pattern so the table cannot disagree with what the rule matches.
file_label <- function(path, pattern) {
  file <- gsub("\\\\", "", sub("\\$$", "", sub("^\\^", "", pattern)))
  ifelse(path == ".", file, paste0(path, "/", file))
}

# Render a markdown table, one row per line, escaping any cell content that
# would otherwise be read as a column separator.
emit_table <- function(headers, ...) {
  cells <- lapply(list(...), function(x) gsub("|", "\\|", x, fixed = TRUE))
  cat(
    c(
      paste0("| ", paste(headers, collapse = " | "), " |"),
      paste0("|", strrep("---|", length(headers))),
      paste0("| ", do.call(paste, c(cells, sep = " | ")), " |")
    ),
    sep = "\n"
  )
  cat("\n\n")
}
```

```{r labels}
# What each code-context and pattern rule covers, in the terms a reader would
# recognize. These cannot be derived from the rule: an XPath expression is a
# definition, not a summary of one. Every rule in the database must appear
# here, so a new rule fails this document until it is described.
code_labels <- c(
  LastLib_base  = "`.Last.lib()`",
  onAttach_base = "`.onAttach()`",
  onDetach_base = "`.onDetach()`",
  onLoad_base   = "`.onLoad()`",
  on_load_rlang = "`rlang::on_load()`",
  onUnload_base = "`.onUnload()`"
)

pattern_labels <- c(
  curl = paste(
    "`curl()`, `curl_fetch_memory()`, `curl_fetch_disk()`,",
    "`curl_fetch_stream()`, `curl_fetch_multi()`, `curl_download()`,",
    "`curl_upload()`, `multi_download()`, `multi_run()`, `send_mail()`"
  ),
  decoding = paste(
    "`base64decode()`, `base64_dec()`, `base64_decode()`, `memDecompress()`,",
    "`rawToChar(as.raw())`"
  ),
  deserialization =
    "`readRDS()`, `load()`, `unserialize()`, `dget()`",
  download_file = "`download.file()`, `url()`",
  dynload       = "`dyn.load()`, `library.dynam()`",
  eval_parse = paste(
    "`eval()`/`evalq()` over `parse()`, `str2lang()`, or `str2expression()`,",
    "when combined with a decoding call"
  ),
  httr = paste(
    "`GET()`, `POST()`, `PUT()`, `PATCH()`, `DELETE()`, `HEAD()`, `VERB()`"
  ),
  httr2 = paste(
    "`req_perform()`, `req_perform_iterative()`, `req_perform_parallel()`,",
    "`req_perform_sequential()`, `req_perform_stream()`,",
    "`req_perform_connection()`, `req_perform_promise()`"
  ),
  indirection = paste(
    "`do.call()`, `match.fun()`, `getFromNamespace()`, `getExportedValue()`,",
    "`getAnywhere()`, `getFunction()`, when the first argument is a string",
    "literal"
  ),
  install = paste(
    "`install.packages()`, `install_github()`, `install_gitlab()`,",
    "`install_bitbucket()`, `install_git()`, `install_url()`,",
    "`install_version()`, `install_local()`, `install_bioc()`,",
    "`install_dev()`, `install_cran()`, `pak()`, `pkg_install()`,",
    "`local_install()`, `lockfile_install()`, `BiocManager::install()`,",
    "`renv::install()`, `devtools::install()`"
  ),
  options_repos = "`options(repos = )`",
  rcurl = paste(
    "`getURL()`, `getURI()`, `getForm()`, `postForm()`, `curlPerform()`"
  ),
  socket = paste(
    "`socketConnection()`, `make.socket()`, `serverSocket()`,",
    "`socketAccept()`"
  ),
  source = "`source()`",
  system_callr = paste(
    "`callr::r()`, `r_bg()`, `r_safe()`, `r_copycat()`, `r_vanilla()`,",
    "`callr::rcmd()`, `rcmd_bg()`, `rcmd_safe()`, `rcmd_copycat()`,",
    "`rscript()`, `r_session$new()`, `r_process$new()`,",
    "`rcmd_process$new()`, `rscript_process$new()`"
  ),
  system_processx = paste(
    "`processx::run()`, `processx::pipeline()`, `process$new()`"
  ),
  system_sys = paste(
    "`exec_wait()`, `exec_background()`, `exec_internal()`, `eval_safe()`,",
    "`eval_fork()`, `r_background()`, `r_internal()`, `r_wait()`"
  ),
  system = "`system()`, `system2()`, `shell()`, `pipe()`"
)

# Descriptions of the two computed contexts. These are authored under
# inst/rules/phases/, which carries phases but no message.
computed_contexts <- data.frame(
  context = c("Top-level", "Other"),
  description = c(
    paste(
      "Top-level code in an R script is evaluated once, when the lazy-load",
      "database is built during installation from source, which R CMD check",
      "and R CMD build also perform. It is not re-evaluated when the namespace",
      "is loaded: loading restores the values from that database."
    ),
    paste(
      "Code inside an ordinary function definition runs at no lifecycle phase.",
      "It executes only if something calls that function, never as a",
      "consequence of building, installing, checking, or loading the package."
    )
  ),
  stringsAsFactors = FALSE
)

# Every rule and computed context must be described above, and every file
# context's derived file name must be matched by the pattern it came from.
stopifnot(
  all(c(rules$file_contexts$name, rules$code_contexts$name,
        rules$patterns$name) %in% names(rule_files)),
  setequal(names(code_labels), rules$code_contexts$name),
  setequal(names(pattern_labels), rules$patterns$name),
  setequal(computed_contexts$context,
           setdiff(rules$phases$context,
                   c(rules$file_contexts$name, rules$code_contexts$name))),
  mapply(
    grepl,
    rules$file_contexts$pattern,
    basename(file_label(rules$file_contexts$path, rules$file_contexts$pattern))
  )
)
```

pkgaudit v`r utils::packageVersion("pkgaudit")` separates *when* code executes
from *what* code does using three categories of rules: file contexts, code
contexts, and patterns. Each rule is defined in a YAML file under
[inst/rules/](https://github.com/tylerjssmith/pkgaudit/tree/master/inst/rules)
and compiled into the SQLite database at `inst/db/rules.db`. The Rule columns
below link to the defining YAML files.

This vignette is generated from that database, at rules
v`r rules_version()`.

## Lifecycle Phases

Every file and code context declares the phases of the package lifecycle in
which its code runs. Each finding carries one logical column per phase, so
findings can be filtered by when they execute, e.g.
`subset(result$patterns, at_install_src)`.

```{r phases-legend}
phase_descriptions <- c(
  at_autoconf    = "Autoconf is run to generate `configure` from its input",
  at_build       = "`R CMD build`",
  at_check       = "`R CMD check`",
  at_install_src = "`R CMD INSTALL` from source",
  at_install_bin = "a prebuilt binary package is installed",
  at_load        = "the namespace is loaded",
  at_attach      = "the package is attached to the search path",
  at_unload      = "the namespace is unloaded",
  at_detach      = "the package is detached from the search path"
)
stopifnot(setequal(names(phase_descriptions), phase_names))

emit_table(
  c("Phase", "Code runs when"),
  paste0("`", phase_names, "`"),
  phase_descriptions[phase_names]
)
```

Phase assignments were established by running `R CMD build`, `R CMD check`, and
`R CMD INSTALL` against instrumented packages rather than read from
documentation. A rule can belong to several phases, and the Phases column below
reads `none` for code that runs at no phase at all.

## File Contexts

File contexts are files that R executes at build-, check-, or install-time.

```{r file-contexts}
fc <- rules$file_contexts
emit_table(
  c("Rule", "File Context", "Phases", "Description"),
  rule_link(fc$name),
  paste0("`", file_label(fc$path, fc$pattern), "`"),
  vapply(fc$name, phase_cell, character(1L)),
  fc$message
)
```

## Code Contexts

Code contexts are lifecycle hooks whose bodies run automatically when a
namespace is loaded, attached, unloaded, or detached.

```{r code-contexts}
cc <- rules$code_contexts
emit_table(
  c("Rule", "Code Context", "Phases", "Description"),
  rule_link(cc$name),
  code_labels[cc$name],
  vapply(cc$name, phase_cell, character(1L)),
  cc$message
)
```

## Computed Contexts

Every pattern is attributed to the code context that contains it. Two of those
contexts are computed rather than matched by a rule, and so have phases but no
rule of their own. They are defined under
[inst/rules/phases/](https://github.com/tylerjssmith/pkgaudit/tree/master/inst/rules/phases).

```{r computed-contexts}
emit_table(
  c("Context", "Phases", "Description"),
  paste0("`", computed_contexts$context, "`"),
  vapply(computed_contexts$context, phase_cell, character(1L)),
  computed_contexts$description
)
```

## Patterns

Patterns are security-relevant function calls. Each pattern finding is
attributed to the code context it executes in, so a `system()` call inside
`.onLoad` is distinguished from one inside an ordinary function ("Other") or at
top level ("Top-level"). Qualified (`pkg::fn()`) and unqualified (`fn()`) call
forms are both detected. Pattern rules carry
[MITRE ATT&CK](https://attack.mitre.org/) technique labels.

A pattern rule declares no phases of its own: a pattern inherits them from the
code context it sits in, listed in the two tables above.

```{r patterns}
pat <- rules$patterns
emit_table(
  c("Rule", "Pattern", "Description"),
  rule_link(pat$name),
  pattern_labels[pat$name],
  pat$message
)
```
