---
title: "pkgaudit Rule Coverage"
output:
  rmarkdown::html_vignette:
    # Rule messages quote shell flags and make constructs verbatim, so the
    # "smart" typography that would turn --clean into an en dash, or
    # $(shell ...) into an ellipsis, is disabled.
    md_extensions: "-smart"
vignette: >
  %\VignetteIndexEntry{pkgaudit Rule Coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "asis")

library(pkgaudit)

# Read the installed database so the tables below cannot drift from the rules
# the package actually ships.
rules       <- load_rules()
phase_names <- setdiff(names(rules$phases), c("context", "version"))
```

```{r helpers}
# The rule sources are read from the installed package, but linked in the
# repository, where a reader can browse them.
repo <- "https://github.com/tylerjssmith/pkgaudit/blob/main/"

# Map each rule name to the YAML file that defines it, as a path relative to
# inst/rules/. The database records a rule's name but not where it came from,
# so the rule sources are read for it.
#
# Keyed by category as well as name: a rule name is unique only within its
# category, and `curl` names both a pattern rule and a match rule.
rule_files <- local({
  root  <- system.file("rules", package = "pkgaudit")
  paths <- list.files(root, pattern = "\\.yaml$", recursive = TRUE)
  named <- vapply(paths, function(path) {
    line <- grep("^name: ", readLines(file.path(root, path)), value = TRUE)
    if (length(line) == 0L) NA_character_ else sub("^name: ", "", line[[1L]])
  }, character(1L))
  # The computed contexts have no rule file to link to, and
  # carry no name field.
  keep <- !is.na(named)
  split(stats::setNames(paths[keep], named[keep]), dirname(paths[keep]))
})

# A rule name, linked to the YAML that defines it in the given category.
rule_link <- function(name, category) {
  sprintf("[%s](%sinst/rules/%s)", name, repo, rule_files[[category]][name])
}

# The phases a context runs in, named as the columns a caller would filter on.
# "none" is what `summary()` calls a context that runs at no phase at all.
phase_cell <- function(context) {
  row <- rules$phases[match(context, rules$phases$context), phase_names]
  on  <- phase_names[as.logical(row)]
  if (length(on) == 0L) "none" else paste0("`", on, "`", collapse = ", ")
}

# A file context's file name, recovered from the rule's search path and
# pattern so the table cannot disagree with what the rule matches.
#
# A rule anchored at both ends names one file. A rule matching a class of files
# -- every R source in R/, say -- has no single name to show, so its pattern is
# shown after a wildcard instead.
anchored <- function(pattern) grepl("^\\^", pattern) & grepl("\\$$", pattern)

file_label <- function(path, pattern) {
  literal <- gsub("\\\\", "", sub("\\$$", "", sub("^\\^", "", pattern)))
  file    <- ifelse(anchored(pattern), literal, paste0("*", literal))
  ifelse(path == ".", file, paste0(path, "/", file))
}

# Render a markdown table, one row per line, escaping any cell content that
# would otherwise be read as a column separator.
emit_table <- function(headers, ...) {
  cells <- lapply(list(...), function(x) gsub("|", "\\|", x, fixed = TRUE))
  cat(
    c(
      paste0("| ", paste(headers, collapse = " | "), " |"),
      paste0("|", strrep("---|", length(headers))),
      paste0("| ", do.call(paste, c(cells, sep = " | ")), " |")
    ),
    sep = "\n"
  )
  cat("\n\n")
}
```

```{r labels}
# What each code-context and pattern rule covers, in the terms a reader would
# recognize. These cannot be derived from the rule: an XPath expression is a
# definition, not a summary of one. Every rule in the database must appear
# here, so a new rule fails this document until it is described.
code_labels <- c(
  LastLib_base     = "`.Last.lib()`",
  onAttach_base    = "`.onAttach()`",
  onDetach_base    = "`.onDetach()`",
  onLoad_base      = "`.onLoad()`",
  on_load_rlang    = "`rlang::on_load()`",
  onUnload_base    = "`.onUnload()`",
  Rd_examples      = "`\\examples{}`",
  Rd_Sexpr_build   = "`\\Sexpr[stage=build]`",
  Rd_Sexpr_install = "`\\Sexpr[stage=install]`",
  Rd_Sexpr_render  = "`\\Sexpr[stage=render]`"
)

pattern_labels <- c(
  chmod = "`Sys.chmod()`, `Sys.umask()`",
  credentials = paste(
    "a path or environment variable that only holds a secret: `~/.ssh/`,",
    "`~/.aws/`, `.netrc`, `id_rsa`, `~/.gnupg/`, and `Sys.getenv()` of a name",
    "containing `TOKEN`, `SECRET`, `PASSWORD` or `API_KEY`"
  ),
  curl = paste(
    "`curl()`, `curl_fetch_memory()`, `curl_fetch_disk()`,",
    "`curl_fetch_stream()`, `curl_fetch_multi()`, `curl_download()`,",
    "`curl_upload()`, `multi_download()`, `multi_run()`, `send_mail()`"
  ),
  decoding = paste(
    "`base64decode()`, `base64_dec()`, `base64_decode()`, `memDecompress()`,",
    "`rawToChar(as.raw())`"
  ),
  deserialization =
    "`readRDS()`, `load()`, `unserialize()`, `dget()`",
  download_file = "`download.file()`, `url()`",
  dynload       = "`dyn.load()`, `library.dynam()`",
  eval_parse = paste(
    "`eval()`/`evalq()` over `parse()`, `str2lang()` or `str2expression()`,",
    "when the text came from a fetch, a decode, a deserialization, a",
    "subprocess or a file read -- every function the rules for those already",
    "cover, plus `readLines()`, `scan()`, `readBin()` and `rawToChar()`"
  ),
  fs   = "`file_chmod()`, `file_chown()`",
  httr = paste(
    "`GET()`, `POST()`, `PUT()`, `PATCH()`, `DELETE()`, `HEAD()`, `VERB()`"
  ),
  httr2 = paste(
    "`req_perform()`, `req_perform_iterative()`, `req_perform_parallel()`,",
    "`req_perform_sequential()`, `req_perform_stream()`,",
    "`req_perform_connection()`, `req_perform_promise()`"
  ),
  install = paste(
    "`install.packages()`, `install_github()`, `install_gitlab()`,",
    "`install_bitbucket()`, `install_git()`, `install_url()`,",
    "`install_version()`, `install_local()`, `install_bioc()`,",
    "`install_dev()`, `install_cran()`, `pak()`, `pkg_install()`,",
    "`local_install()`, `lockfile_install()`, `BiocManager::install()`,",
    "`renv::install()`, `devtools::install()`, `py_install()`,",
    "`virtualenv_install()`, `conda_install()`, `virtualenv_create()`,",
    "`conda_create()`"
  ),
  namespace = paste(
    "`getFromNamespace()`, `getExportedValue()`, `getAnywhere()`,",
    "`assignInNamespace()`, `assignInMyNamespace()`, `unlockBinding()`"
  ),
  options_repos = "`options(repos = )`",
  python = paste(
    "`py_run_string()`, `py_run_file()`, `py_eval()`, `source_python()`,",
    "`py_call()`, `import_builtins()`"
  ),
  persistence = paste(
    "a startup file or scheduler: `~/.bashrc`, `~/.zshrc`, `~/.profile`,",
    "`~/.Rprofile`, `~/.Renviron`, `crontab`, `LaunchAgents`, `/etc/systemd`"
  ),
  rcurl = paste(
    "`getURL()`, `getURI()`, `getForm()`, `postForm()`, `curlPerform()`"
  ),
  socket = paste(
    "`socketConnection()`, `make.socket()`, `serverSocket()`,",
    "`socketAccept()`"
  ),
  source = "`source()`",
  system_callr = paste(
    "`callr::r()`, `r_bg()`, `r_safe()`, `r_copycat()`, `r_vanilla()`,",
    "`callr::rcmd()`, `rcmd_bg()`, `rcmd_safe()`, `rcmd_copycat()`,",
    "`rscript()`, `r_session$new()`, `r_process$new()`,",
    "`rcmd_process$new()`, `rscript_process$new()`"
  ),
  system_processx = paste(
    "`processx::run()`, `processx::pipeline()`, `process$new()`"
  ),
  system_sys = paste(
    "`exec_wait()`, `exec_background()`, `exec_internal()`, `eval_safe()`,",
    "`eval_fork()`, `r_background()`, `r_internal()`, `r_wait()`"
  ),
  system = "`system()`, `system2()`, `shell()`, `pipe()`"
)

# The two computed code contexts, which no rule defines and which carry no
# phases of their own.
computed_contexts <- data.frame(
  context = c("top_level", "in_function"),
  description = c(
    paste(
      "Code outside any function definition. It carries the phases of the file",
      "context it sits in, so the same call reads at_check under tests/ and",
      "at_build under data/."
    ),
    paste(
      "Code inside a function definition. It inherits the phases of the code",
      "around it, except under R/, where the rules report it as running at no",
      "phase."
    )
  ),
  stringsAsFactors = FALSE
)

# Every rule and computed context must be described above, and every file
# context's derived file name must be matched by the pattern it came from.
stopifnot(
  all(rules$file_contexts$name %in% names(rule_files$file_contexts)),
  all(rules$code_contexts$name %in% names(rule_files$code_contexts)),
  all(rules$patterns$name      %in% names(rule_files$patterns)),
  all(rules$matches$name         %in% names(rule_files$matches)),
  setequal(names(code_labels), rules$code_contexts$name),
  setequal(names(pattern_labels), rules$patterns$name),
  # The computed contexts are exactly those with no phases row: they inherit.
  setequal(computed_contexts$context, c("top_level", "in_function")),
  length(setdiff(rules$phases$context,
                 c(rules$file_contexts$name, rules$code_contexts$name))) == 0L,
  # Every code context a file-context rule names must be a rule that exists.
  all(setdiff(unlist(strsplit(
        stats::na.omit(rules$file_contexts$code_context), "[[:space:]]+")),
      "computed") %in% rules$code_contexts$name),
  # Only a rule naming one file has a name to check against its own pattern.
  local({
    fc  <- rules$file_contexts[anchored(rules$file_contexts$filename), ]
    all(mapply(grepl, fc$filename, basename(file_label(fc$path, fc$filename))))
  })
)
```

pkgaudit v`r utils::packageVersion("pkgaudit")` separates *what* code does from
*when* it executes. **Patterns** and **matches** answer the first, and come
first below; **file contexts** and **code contexts** answer the second. Each
rule is defined in a YAML file under
[inst/rules/](https://github.com/tylerjssmith/pkgaudit/tree/main/inst/rules)
and compiled into the SQLite database at `inst/db/rules.db`. The Rule columns
below link to the defining YAML files.

This vignette is generated from that database, at rules
v`r rules_version()`. Every rule in it must be described here, so a rule that
ships without a description fails the build rather than appearing unexplained.

## Patterns

Patterns are security-relevant function calls. Each pattern finding is
attributed to the code context it executes in, so a `system()` call inside
`.onLoad` is distinguished from one inside an ordinary function
(`in_function`) or at top level (`top_level`). Qualified (`pkg::fn()`) and unqualified (`fn()`) call forms are
both detected. Pattern rules carry
[MITRE ATT&CK](https://attack.mitre.org/) technique labels.

A finding also records how the code is *reached*. `guarded` is `TRUE` for code
that ships but the lifecycle does not run -- a `\dontrun{}` block, or a vignette
chunk suppressed by `eval=FALSE` or `#| eval: false`; its phases still come from
its context, so they read as an upper bound. A `\donttest{}` block is not guarded:
a plain `R CMD check` skips it, but `--as-cran` runs it. `indirect` is `TRUE`
where the call was made through the function's name rather than the function.

A pattern rule declares no phases of its own. A pattern takes them from the file
context it was found in and the code context it sits in, resolved as
[Lifecycle Phases](#lifecycle-phases) describes.

A call made through a function's name rather than the function --
`do.call("system", ...)`, `match.fun()`, `getFunction()` -- is reported under
the rule that owns the name, with the `indirect` column set. Each rule declares
the names it claims, and a name is only accepted if calling it bare would have
matched that same rule, so an indirect finding can never be attributed to a
rule that would not have reported the direct call.

```{r nameless, echo = FALSE}
fns      <- rules$patterns$functions
nameless <- sort(rules$patterns$name[is.na(fns) | !nzchar(fns)])
```

`r length(nameless)` rules claim no names -- `r knitr::combine_words(sprintf("\x60%s\x60", nameless))` --
for two different reasons. `eval_parse`, `options_repos` and `system_processx`
match on more than the callee: a nested call, a named argument, a package
qualifier, none of which survives the trip through `do.call()`. `credentials`
and `persistence` match string constants rather than calls, so there is no
callee to claim in the first place. A name assembled at runtime is not resolved
either, since pkgaudit evaluates nothing.

```{r patterns}
pat <- rules$patterns
emit_table(
  c("Rule", "Pattern", "Description"),
  rule_link(pat$name, "patterns"),
  pattern_labels[pat$name],
  pat$message
)
```

## Matches

Matches are regular-expression matches in the shell scripts and Make-like
files among the [file contexts](#file-contexts) below -- those whose Type is
`shell` or `make`.
A file context typed `R` is parsed and scanned for patterns instead, and no
other file in the package is scanned for matches at all. Match rules
carry [MITRE ATT&CK](https://attack.mitre.org/) technique labels.

Like a pattern rule, a match rule declares no phases of its own. Where a file
matches more than one file-context rule, it takes the phases of every rule that
matched it.

Matching text is less precise than matching a parse tree. A match has no
syntax behind it, so a match inside a comment, a quoted string, or a branch that
never runs is reported the same as one in a live command. A match is a candidate
for review rather than confirmed behaviour, and warrants reading the file itself
more than a pattern finding does.

```{r matches}
rx <- rules$matches
emit_table(
  c("Rule", "Description"),
  rule_link(rx$name, "matches"),
  rx$message
)
```

The regular expressions themselves are matched with `perl = TRUE` and are
case-sensitive:

```{r matches-regex, results = "markup", comment = ""}
# Shown as a verbatim block rather than a table column: an alternation `|` in a
# regular expression would be read as a column separator, and escaping it
# leaves the escape visible in the rendered cell.
cat(
  sprintf("%s  %s", formatC(rx$name, width = max(nchar(rx$name)), flag = "-"),
          rx$regex),
  sep = "\n"
)
```

## Lifecycle Phases

A phase is a property of two things: where a file sits in the package, and where
the code sits within that file. Each finding carries one logical column per
phase, so findings can be filtered by when they execute, e.g.
`subset(result$patterns, at_install_src)`.

```{r phases-legend}
phase_descriptions <- c(
  at_autoconf    = "Autoconf is run to generate `configure` from its input",
  at_build       = "`R CMD build`",
  at_check       = "`R CMD check`",
  at_install_src = "`R CMD INSTALL` from source",
  at_install_bin = "a prebuilt binary package is installed",
  at_load        = "the namespace is loaded",
  at_attach      = "the package is attached to the search path",
  at_unload      = "the namespace is unloaded",
  at_detach      = "the package is detached from the search path"
)
stopifnot(setequal(names(phase_descriptions), phase_names))

emit_table(
  c("Phase", "Code runs when"),
  paste0("`", phase_names, "`"),
  phase_descriptions[phase_names]
)
```

A match takes the phases of the file context it was found in. A pattern resolves
them in order: a code context carrying phases of its own wins, and otherwise the
code inherits the phases of the file context around it. So the same `system()`
call reads `at_check` under `tests/` and `at_build` under `data/`.

Code inside a function definition inherits too, unless its file's rule sets
`assume_called: FALSE`, in which case it is reported as running at no phase.
Only the rules for `R/` set it. Both readings are measured -- a function called
from top-level code runs whenever that code does, and one nothing calls runs
nowhere -- so a rule is choosing which measurement to report, and cannot state a
third. Neither is a claim about any particular package's call graph, which
pkgaudit does not trace.

Every phase value below was established by running instrumented packages through
`R CMD build`, `R CMD check` and `R CMD INSTALL` and recording which sites fired,
rather than read from documentation. A rule can belong to several phases, and a
Phases column reads `none` for code that runs at no phase at all.

## File Contexts

File contexts are files that R executes at build-, check-, or install-time.

The Type column determines how a matched file is read. Files typed `R` are
parsed and scanned for patterns; `Rd` files have their `\examples{}` and
`\Sexpr{}` code extracted and scanned for patterns; `shell` and `make` files are
matched against the rules listed under [Matches](#matches).

The Report column separates being scanned from being reported. Every rule tells
the scan which files to read; only a reporting rule contributes a row to the
`file_contexts` findings frame.

`report` is `TRUE` for a file that meets two conditions: it executes
automatically during at least one lifecycle phase, and its contents are matched
as text rather than parsed. Together those mean pkgaudit cannot tell a reviewer
what the file does, only that it runs -- so the file itself is the finding, and
someone has to read it. In practice that is the shell scripts and Make-like
files: `configure`, `cleanup`, `src/Makevars` and their variants.

Everything else is scanned just as thoroughly. A pattern in a vignette or a test
file is reported with its context and phases whether or not the file itself is.
What `report` withholds is the row asserting the file exists -- which, for a file
whose contents are parsed, the findings already tell you.

`src/install.libs.R` is the one exception: it is R, and parsed, but its presence
replaces R's default handling of compiled artifacts, a structural change that
follows from the file existing rather than from anything written in it.

```{r file-contexts}
fc <- rules$file_contexts
emit_table(
  c("Rule", "File Context", "Type", "Report", "Phases", "Description"),
  rule_link(fc$name, "file_contexts"),
  paste0("`", file_label(fc$path, fc$filename), "`"),
  paste0("`", fc$type, "`"),
  ifelse(fc$report, "yes", "no"),
  vapply(fc$name, phase_cell, character(1L)),
  fc$message
)
```

## Code Contexts

A code context says where code sits *within* its file. Every pattern carries
one, and it is half of what decides the finding's phases.

Two kinds are rules. A **lifecycle hook** is found by matching the parse tree,
and applies only where a package's R code becomes its namespace: `R/`,
`R/unix/`, `R/windows/`. A `.onLoad` defined anywhere else ships as an ordinary
object and never fires -- measured, not assumed. A **part of a help file** is
found by matching a label the extractor stamped, because the distinction is
invisible to the parse tree: once an `.Rd` has yielded R code, nothing in that
code says whether it came from `\examples` or from a `\Sexpr`, or which stage.

Both carry phases of their own, so a finding in one takes them directly.

```{r code-contexts}
cc <- rules$code_contexts
emit_table(
  c("Rule", "Code Context", "Matched by", "Phases", "Description"),
  rule_link(cc$name, "code_contexts"),
  code_labels[cc$name],
  ifelse(cc$kind == "xpath", "parse tree", "segment label"),
  vapply(cc$name, phase_cell, character(1L)),
  cc$message
)
```

The remaining two are computed from the parse tree and are the only contexts
with no phases of their own. They inherit from the file context around them,
which is why there is no `data` or `tests` code context: a `system()` call at
the top level of a data script and one in a test file sit in the same place in
their files, and what differs is when the file runs.

```{r computed-contexts}
emit_table(
  c("Context", "Phases", "Description"),
  paste0("`", computed_contexts$context, "`"),
  rep("inherited", nrow(computed_contexts)),
  computed_contexts$description
)
```
