---
title: "Understanding pkgaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding pkgaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit is a static analysis security testing (SAST) tool for R source
packages. It flags security-relevant files and R source code for manual review,
and organizes what it finds by the point in the R package lifecycle at which the
code runs (e.g., build, check, install, load).

This vignette explains how pkgaudit works. It is intended for anyone considering
whether to incorporate pkgaudit into their workflow and for potential
contributors. For information on how to contribute, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/master/.github/CONTRIBUTING.md).

```{r setup}
library(pkgaudit)
rules <- load_rules()
```

## Entry Points

pkgaudit has two entry points. `audit_package()` is used for source package
directories. `audit_tarball()` is used for source package tarballs; it wraps
`audit_package()` and handles tarball validation and extraction, and the removal
of extracted files.

`audit_package()` orchestrates the following internal functions:

1. `find_file_contexts()` determines what file contexts are present in the
    source package.

2. `find_scripts()` finds all scripts in `R/`, `R/unix/`, and `R/windows/` with
    file extensions recognized by R's build process (`.R`, `.r`, `.S`, `.s`,
    `.q`), as well as `src/install.libs.R`, if present.

3. For each script: `parse_script()` uses R's `parse()` and the `xmlparsedata`
   and `xml2` packages to return an XML parse tree for the script.

   a. For each code-context rule, `find_code_contexts()` finds XML nodes
      matching XPath-based definitions of code contexts.

   b. For each pattern rule, `find_patterns()` finds XML nodes matching
      XPath-based definitions of patterns.

   c. If any patterns were found, `determine_code_contexts()` determines
      whether each pattern occurrence is located in a named code context,
      top-level code (`Top-level`), or a regular function (`Other`).

4. Finally, the lifecycle phases in which each finding's code runs are attached
   to the findings. A file or code context takes its phases from the rule that
   matched it; a pattern inherits them from the code context assigned in
   step 3c.

`audit_package()` returns a `pkgaudit` object: a named list of `file_contexts`,
`code_contexts`, `patterns`, and `errors` data frames, plus a `metadata` list
recording what was scanned and with what. Each of the three findings data frames
carries one logical column per lifecycle phase.

Failures are contained rather than fatal. Listing files, parsing a script, and
evaluating an XPath are each wrapped in `tryCatch()`, so an unreadable file, a
malformed script, or an invalid XPath is recorded as a row in `errors` -- giving
the stage, file, rule, and message -- and the scan continues with the next
script or rule.

## Rules

The remainder of this document explains each category of rule and how the
functions above are used to find the context or pattern it targets.

Rules are defined by YAML files that live under `inst/rules/`. For all three
categories, the YAML files have the following fields:

- `name` is the rule identifier and the value reported in a finding.
- `version` is the rule version, independent of the package version.
- `message` is the explanation shown with a finding.
- `positive_examples` and `negative_examples` are file paths or code syntax that
should and should not be found by the rule, respectively.

File and code context rules carry nine further fields, one per lifecycle phase
-- `at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`on_load`, `on_attach`, `on_unload`, and `on_detach` -- each `TRUE` or `FALSE`.
Pattern rules do not declare them: a pattern runs when the code around it runs,
so it takes its phases from the code context that contains it.

Each rule contains additional fields used by pkgaudit functions to find file
contexts, code contexts, or patterns. These are explained below.

The helper below prints a rule's YAML, stopping before its examples, which are
discussed at the end.

```{r helper}
show_rule <- function(category, file) {
  lines <- readLines(system.file("rules", category, file, package = "pkgaudit"))
  end   <- grep("^positive_examples:", lines)
  cat(lines[seq_len(if (length(end)) end[1] - 1L else length(lines))], sep = "\n")
}
```

### File Context Rules

A file context is a file that R executes at build-, check-, or install-time,
such as `configure` and `src/Makevars`. These files have many legitimate uses
but can execute arbitrary shell commands on a system. The goal of a file context
rule is to alert a user to the existence of a file context before it is executed
so that the file can be reviewed.

In addition to the common fields and the nine phase fields, a file context rule
has four additional fields:

- `type` is the language or format of the file (e.g., `make`, `shell`, `R`).
- `path` is the directory to search (relative to the package root).
- `recursive` indicates whether to descend into subdirectories.
- `pattern` is a regular expression matched against the names of the files found
  at the file path (recursively, if `recursive: TRUE`).

`find_file_contexts()` passes the `path`, `recursive`, and `pattern` values to
`base::list.files()`. For example, the following rule is used to find
`configure` in the package root.

```{r fc-yaml, comment = ""}
show_rule("file_contexts", "file_configure.yaml")
```

`"."` is the package root. `recursive: FALSE` means search only in the package
root, not subdirectories. `pattern` is anchored at both ends, so `^configure$`
matches `configure` exactly.

```{r fc-match}
grepl("^configure$", c("configure", "configure.win", "reconfigure"))
```

For this rule, these values are passed to `list.files()` as follows, with `path`
resolved against the root of the package being audited:

```{r, eval=FALSE}
list.files(
  file.path(pkg, "."),
  pattern    = "^configure$",
  recursive  = FALSE,
  full.names = TRUE,
  all.files  = TRUE
)
```

All matches are findings. Here at most one file can match, since the pattern is
anchored and the search does not descend into subdirectories. Related files such
as `configure.win` and `configure.ucrt` have separate rules rather than a shared
pattern. Additionally, pkgaudit ignores directories whose names happen to match;
only regular files are found.

The YAML files are compiled into a SQLite database. `load_rules()` returns a
list of four data frames: one per rule category, plus `phases`. The
`file_contexts` data frame contains the `path`, `recursive`, and `pattern`
values:

```{r fc-row}
rules$file_contexts[
  rules$file_contexts$name == "configure_file",
  c("name", "path", "recursive", "pattern")
]
```

The rule's phase fields are hoisted into `phases`, keyed by rule name. Here they
say that `configure` runs during `R CMD build` and `R CMD check` as well as at
installation from source, and never once the package is loaded:

```{r fc-phases}
rules$phases[
  rules$phases$context == "configure_file",
  c("context", "at_build", "at_check", "at_install_src", "on_load")
]
```

### Code Context Rules

A code context is a lifecycle hook whose body can run automatically when a
namespace is loaded, attached, unloaded, or detached, such as `.onLoad()` and
`.onAttach()`. These contexts have many legitimate uses but can execute
arbitrary code on a system when a user installs or loads a package, before the
user has called any package functions. The goal of a code context rule is to
alert a user to the existence of a code context before its contents are executed
so that its body can be reviewed.

For each script found by `find_scripts()`, `parse_script()` returns an XML parse
tree. As a simplified example, the helpers below are used to parse a minimal
example of `.onLoad()` and print its tree, dropping the line and column
attributes to keep the output readable.

```{r cc-helpers}
as_xml <- function(code) {
  xml2::read_xml(
    xmlparsedata::xml_parse_data(
      parse(text = code, keep.source = TRUE)
    )
  )
}

show_tree <- function(code) {
  doc <- as_xml(code)
  for (node in xml2::xml_find_all(doc, "//*")) {
    for (attr in names(xml2::xml_attrs(node))) xml2::xml_attr(node, attr) <- NULL
  }
  cat(sub("^<\\?xml[^>]*\\?>\n", "", as.character(doc)))
}
```

```{r cc-tree, comment = ""}
show_tree(".onLoad <- function(libname, pkgname) NULL")
```

In addition to the common fields and the nine phase fields, a code context rule
has two additional fields:

- `type` is the language of the code the context is written in (`R`).
- `xpath` is the XPath expression that defines the context.

```{r cc-yaml, comment = ""}
show_rule("code_contexts", "code_onload.yaml")
```

The XPath definition selects an expression node subject to two alternative
conditions. The first covers assignment with `<-` or `=`: the left side must be
the symbol `.onLoad`, and the right side must be a function definition, which
is what `expr[FUNCTION]` tests. The second covers the same definition written
backwards with `->`, where the symbol and the function swap sides. Expressions
are matched under two node names because R wraps a top-level assignment made
with `=` in an `expr_or_assign_or_help` node rather than an `expr` node.

`find_code_contexts()` takes the XML parse tree and XPath definition, and calls
`.xml_find_all_safe()`, which wraps `xml2::xml_find_all()`, to find XML nodes in
the parse tree matching the definition. Every matching node becomes a
code-context finding recording the rule's name and the line and column where the
hook begins.

Importantly, for the definition to match, a *function* must be assigned to the
name of the hook. The name of the hook merely appearing somewhere is not a
match. Below are some simple examples:

```{r cc-match}
xpath   <- rules$code_contexts$xpath[rules$code_contexts$name == "onload_code"]
matches <- function(code) length(xml2::xml_find_all(as_xml(code), xpath)) > 0

matches(".onLoad <- function(libname, pkgname) NULL")    # assignment
matches(".onLoad = function(libname, pkgname) NULL")     # assignment with `=`
matches("(function(libname, pkgname) NULL) -> .onLoad")  # right assignment
matches(".onAttach <- function(libname, pkgname) NULL")  # a different hook
matches("message('.onLoad')")                            # only a mention
```

### Pattern Rules

A pattern is a security-relevant function call, such as `system()`, which can
execute arbitrary shell commands. These functions have many legitimate uses. The
goal of a pattern rule is to alert a user to the existence of a function call
that should be reviewed.

In addition to the common fields, a pattern rule has three additional fields:

- `type` is the language of the code the pattern matches (`R`).
- `xpath` is the XPath expression that defines the pattern.
- `attck` is a set of MITRE ATT&CK technique identifiers, carried into each
  finding and tallied by `summary()`.

A pattern rule carries no severity. How much a finding deserves your attention
depends on the pattern together with the context it was found in -- the same
`system()` call is weightier in `.onLoad()` than in a function nothing calls --
and a rule is evaluated without knowing which context it will land in.

As with code contexts, `find_patterns()` takes the XML parse tree and XPath
definition, and calls `.xml_find_all_safe()`, which wraps
`xml2::xml_find_all()`, to find XML nodes in the parse tree matching the
definition. Every matching node becomes a pattern finding recording the rule's
name and the line and column where the call begins.

As a simplified example, the XML parse tree for `system()` is:

```{r system-example, comment = ""}
show_tree("system()")
```

The XPath definition of the pattern rule matching `system()` and similar
function calls is:

```{r pat-yaml, comment = ""}
show_rule("patterns", "pattern_system.yaml")
```

There are two details in this XPath worth noting. The first is
`SYMBOL_FUNCTION_CALL`. R's parser uses that label only where a name is actually
a function call; a mention of the same name anywhere else is a plain `SYMBOL`,
which will not match.

```{r pat-symbol, comment = ""}
show_tree("foo <- system")
```

The second is the pair of `not(...)` conditions. In `foo$system("x")`, the name
is in call position, so the parser labels it `SYMBOL_FUNCTION_CALL` -- but it is
a list element or an object's method, not base R's `system()`. The rule
excludes it by refusing any call whose name is immediately preceded by `$`:

```{r pat-dollar, comment = ""}
show_tree('foo$system("x")')
```

The `@` half of that pair is belt and braces. R's parser labels the name after
`@` as a `SLOT` rather than a call, so `foo@system("x")` will not match, but
the condition is included as future-proofing.

```{r pat-at, comment = ""}
show_tree('foo@system("x")')
```

The following shows examples of what does and does not match:

```{r pat-match}
xpath_sys <- rules$patterns$xpath[rules$patterns$name == "system_pattern"]
hits <- function(code) length(xml2::xml_find_all(as_xml(code), xpath_sys)) > 0

hits('system("id")')
hits('base::system("id")')   # a qualified call still matches
hits('pipe("id")')           # one of the four names the rule covers
hits("foo <- system")        # a reference, not a call
hits('foo$system("id")')     # a different function
```

`base::system("id")` still matches because the name is preceded by the package
qualifier (`::`), not by `$` or `@`. Detecting both call forms is why the rule
tests for the exclusions rather than requiring the call to stand alone.

## Putting Them Together

`determine_code_contexts()` re-evaluates the code context XPath definitions
against the XML parse tree and compares the resulting nodes with the pattern
nodes already found, assigning each pattern the most specific code context that
contains it. A pattern at the top level of a script is assigned `Top-level`, and
a pattern inside a regular function is assigned `Other`.

That assignment is also what gives a pattern its phases. `Top-level` and `Other`
are computed rather than matched by a rule, so they have no rule to take phases
from; theirs are authored separately, under `inst/rules/phases/`. Top-level code
is evaluated once, when the lazy-load database is built during installation from
source, so `Top-level` carries the build, check, and source-install phases.
`Other` carries none at all, which is the reading intended: code inside an
ordinary function runs only if something calls it, never as a consequence of
building, installing, checking, or loading the package.

## Examples Are Tests

Every rule YAML ends with `positive_examples` and `negative_examples`.

```{r examples, comment = ""}
lines <- readLines(
  system.file("rules", "file_contexts", "file_configure.yaml",
    package = "pkgaudit")
)
cat(lines[grep("^positive_examples:", lines):length(lines)], sep = "\n")
```

Every example is written out as a test fixture, and the pkgaudit test suite
requires that every positive example is flagged by its rule and that no negative
example is. A rule whose examples do not behave as written fails the suite, so
the examples stay accurate as rules change.

The full rule set is documented in
[RULES.md](https://github.com/tylerjssmith/pkgaudit/blob/master/RULES.md), and
the rules themselves live under
[inst/rules/](https://github.com/tylerjssmith/pkgaudit/tree/master/inst/rules).
