---
title: "Understanding pkgaudit Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding pkgaudit Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

pkgaudit scans R source packages for security-relevant files and R source code,
and organizes what it finds by the point in the R package lifecycle at which the
code runs (e.g., build, check, install, load).

This scan is based on rules, which are defined by YAML files under
`inst/rules/` and compiled into the SQLite database that ships with the package.
Adding a rule means writing a YAML file, not writing R.

This vignette walks through one rule from each of the three categories (file
contexts, code contexts, and patterns) and shows how its fields are used by
pkgaudit.

```{r setup}
library(pkgaudit)

rules <- load_rules()
```

All three categories share three fields:

- `name` is the rule identifier and the value reported in a finding.
- `version` is the rule version, independent of the package version.
- `message` is the explanation shown with a finding.

Each rule contains additional fields used by pkgaudit functions to find file
contexts, code contexts, or patterns.

The helper below prints a rule's YAML, stopping before its examples, which we
return to at the end.

```{r helper}
show_rule <- function(category, file) {
  lines <- readLines(system.file("rules", category, file, package = "pkgaudit"))
  end   <- grep("^positive_examples:", lines)
  cat(lines[seq_len(if (length(end)) end[1] - 1L else length(lines))], sep = "\n")
}
```

## File Context Rules

A file context rule asks: Does a file with this name exist in this part of the 
package? No R code is read or parsed. This category of rule has four additional
fields:

- `type` is the type of file (e.g., `shell`, `make`).
- `path` is the directory to search, relative to the package root.
- `recursive` indicates whether to descend into subdirectories.
- `pattern` is a regular expression matched against the names of the files found
  there.

The following rule is used to find `configure` in the package root.

```{r fc-yaml, comment = ""}
show_rule("file_contexts", "file_configure.yaml")
```

`"."` is the package root. `recursive: FALSE` means search only in the package
root, not subdirectories. `pattern` is anchored at both ends, so `^configure$`
matches `configure` exactly.

```{r fc-match}
grepl("^configure$", c("configure", "configure.win", "reconfigure"))
```

Related files such as `configure.win` and `configure.ucrt` have separate rules 
rather than a shared pattern. Additionally, pkgaudit ignores directories whose 
names happen to match; only regular files are found.

The compiled rule is a row in a data frame in the list returned by 
`load_rules()`, and these three fields are the whole of its matching logic:

```{r fc-row}
rules$file_contexts[
  rules$file_contexts$name == "configure_file",
  c("name", "path", "recursive", "pattern")
]
```

## Code Context Rules

A code context rule asks: Is a specific lifecycle hook defined in a script? That
depends on the structure of the code rather than its text, so pkgaudit parses
each R file and matches against an XML parse tree using XPath. This category of 
rule has two additional fields:

- `type` is the type of code context (e.g., `hook`).
- `xpath` is the XPath expression used to find the code context.

```{r cc-yaml, comment = ""}
show_rule("code_contexts", "code_onload.yaml")
```

R's parser records the structure of the code it reads, and the `xmlparsedata`
and `xml2` packages turn that record into an XML parse tree. The helpers below 
parse a snippet and print its tree, dropping the line and column attributes to 
keep the output readable.

```{r cc-helpers}
as_xml <- function(code) {
  xml2::read_xml(
    xmlparsedata::xml_parse_data(
      parse(text = code, keep.source = TRUE)
    )
  )
}

show_tree <- function(code) {
  doc <- as_xml(code)
  for (node in xml2::xml_find_all(doc, "//*")) {
    for (attr in names(xml2::xml_attrs(node))) xml2::xml_attr(node, attr) <- NULL
  }
  cat(sub("^<\\?xml[^>]*\\?>\n", "", as.character(doc)))
}
```

Here is what the `.onLoad()` hook looks like as an XML parse tree:

```{r cc-tree, comment = ""}
show_tree(".onLoad <- function(libname, pkgname) NULL")
```

The rule's XPath selects an expression node subject to two alternative
conditions. The first covers assignment with `<-` or `=`: the left side must be
the symbol `.onLoad`, and the right side must be a function definition, which
is what `expr[FUNCTION]` tests. The second covers the same definition written
backwards with `->`, where the symbol and the function swap sides. Expressions
are matched under two node names because R wraps a top-level assignment made
with `=` in an `expr_or_assign_or_help` node rather than an `expr` node.

What the rule does *not* match is the point. It requires a function to be
assigned to that name, not merely that the name appears somewhere:

```{r cc-match}
xpath   <- rules$code_contexts$xpath[rules$code_contexts$name == "onload_code"]
matches <- function(code) length(xml2::xml_find_all(as_xml(code), xpath)) > 0

matches(".onLoad <- function(libname, pkgname) NULL")    # assignment
matches(".onLoad = function(libname, pkgname) NULL")     # assignment with `=`
matches("(function(libname, pkgname) NULL) -> .onLoad")  # right assignment
matches(".onAttach <- function(libname, pkgname) NULL")  # a different hook
matches("message('.onLoad')")                            # only a mention
```

Every matching node becomes a code-context finding recording the rule's name
and the line and column where the hook begins.

## Pattern Rules

A pattern rule asks: Does a script call one of a set of functions? It uses
the same XML parse tree and the same XPath machinery as a code context rule, but
it matches call sites rather than definitions. This category of rule has three
additional fields:

- `type` is a severity (`warning` or `note`).
- `xpath` is the XPath expression used to find the pattern.
- `attck` contains MITRE ATT&CK labels.

```{r pat-yaml, comment = ""}
show_rule("patterns", "pattern_system.yaml")
```

Two details in this XPath carry most of the weight. The first is 
`SYMBOL_FUNCTION_CALL`. R's parser uses that label only where a name is actually
being called; a mention of the same name anywhere else is a plain `SYMBOL`.
Critically, a reference to the function, rather than a call to it, produces a 
tree the rule cannot match:

```{r pat-symbol, comment = ""}
show_tree("foo <- system")
```

The second is the pair of `not(...)` conditions. In `foo$system("x")` the name
is genuinely in call position, so the parser does label it
`SYMBOL_FUNCTION_CALL` -- but it is a list element or an object's method, not
base R's `system()`. The rule excludes it by refusing any call whose name is
immediately preceded by `$`:

```{r pat-dollar, comment = ""}
show_tree('foo$system("x")')
```

The `@` half of that pair is belt and braces. R's parser labels the name after
`@` as a `SLOT` rather than a call, so `foo@system("x")` could never have
matched in the first place:

```{r pat-at, comment = ""}
show_tree('foo@system("x")')
```

Together:

```{r pat-match}
xpath_sys <- rules$patterns$xpath[rules$patterns$name == "system_pattern"]
hits <- function(code) length(xml2::xml_find_all(as_xml(code), xpath_sys)) > 0

hits('system("id")')
hits('base::system("id")')   # a qualified call still matches
hits('pipe("id")')           # one of the four names the rule covers
hits("foo <- system")        # a reference, not a call
hits('foo$system("id")')     # a different function
```

`base::system("id")` still matches because the name is preceded by the package
qualifier (`::`), not by `$` or `@`. Detecting both call forms is why the rule
tests for the exclusions rather than requiring the call to stand alone.

## Putting the Three Together

A pattern match on its own says only that a function call exists somewhere. By 
determining file and code contexts, pkgaudit helps a user understand when the
call may occur. A `system()` call inside `.onLoad()` runs when the namespace
is loaded. The same call inside an ordinary function is reported as `"Other"`
and runs only if a user calls that function. The same call at the top level of a
script is reported as `"Top-level"`, and runs when the namespace is built.

## Examples Are Tests

Every rule YAML ends with `positive_examples` and `negative_examples`, which we
skipped above:

```{r examples, comment = ""}
lines <- readLines(
  system.file("rules", "file_contexts", "file_configure.yaml", 
    package = "pkgaudit")
)
cat(lines[grep("^positive_examples:", lines):length(lines)], sep = "\n")
```

These are not only documentation. Every example is written out as a test
fixture, and the pkgaudit test suite requires that every positive example is
flagged by
its rule and that no negative example is. A rule whose examples do not behave as
written fails the suite, so the examples stay accurate as rules change.

The full rule set is documented in
[RULES.md](https://github.com/tylerjssmith/pkgaudit/blob/master/RULES.md), and
the rules themselves live under
[inst/rules/](https://github.com/tylerjssmith/pkgaudit/tree/master/inst/rules).
