---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit index.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# pkgaudit

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit scans R packages for security-relevant files and code without executing
anything it scans. It reports what code does and when it runs, so code that runs
on install or load is distinguishable from code that runs only when called.

A general-purpose scanner like Semgrep can read R -- but it reads scripts, not
packages. It does not look for R inside an `\examples{}` block, an `\Sexpr{}`
macro, or a vignette. It does not know that `.onLoad()` runs when a user
calls `library()` or that a `configure` script runs on `R CMD check`. pkgaudit
extracts code wherever it exists, scans it for functions and commands that need
human review, and reports the lifecycle phases in which each finding runs.

For why this matters, see [R Package Security](articles/security.html). For the
rule set, see [Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Phases overlap -- building a package with vignettes, for example, also installs
and loads it -- and one occurrence is counted under every phase it runs in. The
summary above reflects five findings, some counted under multiple phases.

pkgaudit can integrate its scan with other tools. `emit_sarif()` renders its
results as SARIF 2.1.0, which editors and code-scanning platforms read directly. `export_unscanned()` exports code written in languages pkgaudit cannot read to a
directory for a scanner that can. 

See [Getting Started with pkgaudit](articles/pkgaudit.html) for details.

## Security

pkgaudit's own security model, and how to report a vulnerability, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
