---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit index.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# pkgaudit

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit scans R packages for security-relevant files and code without executing
anything it scans. It reports what code does and when it runs, so code that runs
on install or load is distinguishable from code that runs only when called.

A general-purpose scanner like Semgrep can read R -- but it reads scripts, not
packages. It does not look for R inside an `\examples{}` block, an `\Sexpr{}`
macro, or a vignette. It does not know that `.onLoad()` runs when a user
calls `library()` or that a `configure` script runs on `R CMD check`. pkgaudit
extracts code wherever it exists, scans it for functions and commands that need
human review, and reports the lifecycle phases in which each finding runs.

For why this matters, see [R Package Security](articles/security.html). For the
rule set, see [Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Phases overlap -- building a package with vignettes, for example, also installs
and loads it -- and one occurrence is counted under every phase it runs in. The
summary above reflects five findings, some counted under multiple phases:

1. An `\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help
page is rendered at build, check, and source installation. Reviewers should
consider whether a help file should make an HTTP POST request and what it can
send to an external host.

2. `.onLoad()` in `R/zzz.R` calls `system()` and runs when users call
`library()`. It also runs at build, check, and source installation, each of
which loads the package. While `system()` calls are common in R packages, they
are less common in lifecycle hooks like `.onLoad()`, and reviewers should
inspect what commands would be run automatically on their systems.

3. The `\examples{}` block in the same help file calls `download.file()` at
check. Reviewers should verify what is downloaded. Since this depends on an
external host, and what is downloaded may change over time, reviewers should
also consider how the download is used -- for example, could another function
execute code if the file ever contained it?

4. A regular function in `R/fetch.R` calls `download.file()`, but a regular
function is not known to run automatically, so this pattern is reported under
`none`. A reviewer may still want to inspect if and when the function is called,
and what would be downloaded.

5. The `configure` script may invoke `curl` at build, check, and source
installation. Some R packages use `curl` to fetch dependencies that cannot be
vendored with the package. Reviewers should verify what is fetched and what
could happen if that changes.

pkgaudit also provides functions to integrate its scan with other tools.
`emit_sarif()` renders its results as SARIF 2.1.0, which editors and
code-scanning platforms read directly. `export_unscanned()` writes out package
code in languages pkgaudit cannot read, into a directory for a scanner that
can. See [Getting Started with pkgaudit](articles/pkgaudit.html) for details.

## Security

pkgaudit's own security model, and how to report a vulnerability, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
