---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit index.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# pkgaudit

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
flags security-relevant files and code in R source packages for human review
without executing anything it scans. pkgaudit also models package lifecycle
execution semantics -- it reports not just what a package does, but when it
runs, so code that executes on install or load is distinguishable from code that
runs only when called.

A general-purpose scanner can read R -- Semgrep supports it experimentally --
but it reads files, not packages. Nothing tells it to look for R inside a help
file's `\examples{}` block or an `\Sexpr{}` macro, or inside a Sweave or R.rsp
vignette; and nothing tells it that `.onLoad()` runs on `library()`, that
`\Sexpr{}` evaluates while a help page is built, or that `configure` runs under
`R CMD check`. pkgaudit does this: it extracts R from wherever a package carries
it, parses the code for security-relevant patterns, and reports the lifecycle
phases in which each finding runs.

For why this matters, see [R Package Security](articles/security.html). For the
rule set, see [Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Four of `untrustedpkg`'s five findings run without anyone asking for them.
`.onLoad()` in `R/zzz.R` calls `system()`, so it runs on `library()` -- and at
build, check, and source installation, each of which loads the package. An
`\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help page
is rendered, at those same three phases, and the `\examples{}` block in the same
file calls `download.file()`, which `R CMD check` runs. The `configure` script
may invoke `curl` at build, check, and source installation. The fifth finding is
the contrast that makes the rest legible: `download.file()` in `R/fetch.R` sits
in an ordinary function body, so it runs at no phase at all, reported under
`none`, because it executes only if someone calls it.

Two functions carry a scan into other tools. `emit_sarif()` renders the result
as SARIF 2.1.0, which editors and code-scanning platforms read directly.
`export_unscanned()` writes the code pkgaudit cannot read into a directory for
a scanner that can. See [Getting Started with pkgaudit](articles/pkgaudit.html)
for details.

## Security

pkgaudit's own security model, and how to report a vulnerability in it, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
