---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Published SHA-256 of the shipped rules database. Read from the repository copy
# so the value below cannot drift from the database actually committed. knitr
# makes this file's own directory the working directory, so the repository root
# is one level up; knitting from the root is tolerated as well.
db_path <- file.path("..", "inst", "db", "rules.db")
if (!file.exists(db_path)) db_path <- file.path("inst", "db", "rules.db")
stopifnot(file.exists(db_path))
db_hash <- digest::digest(db_path, algo = "sha256", file = TRUE)

# Wrap in backticks for inline use. The backtick is built from its code point so
# that it does not terminate the inline expression that emits it.
backtick    <- intToUtf8(96L)
db_hash_md  <- paste0(backtick, db_hash, backtick)
```

# pkgaudit

[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![test coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
reports which parts of a package can execute, when they execute, and what they
do, so that an untrusted package can be reviewed before it is installed and
loaded. It never executes the code it scans.

## How a package is covered

pkgaudit accounts for every file it can identify as code, and says what it made
of each one:

- **parsed** -- R, wherever a package carries it: `R/`, help-file examples and
  `\Sexpr{}` macros, vignettes in R Markdown, Quarto, Sweave and R.rsp,
  `data/`, `demo/`, `tests/`, `tools/`, `inst/CITATION`, `.Rprofile`. Matched
  against R's parse tree.
- **matched** -- shell scripts and Make-like files: `configure`, `cleanup`,
  `src/Makevars`. Matched as text, which is less precise -- a match in a
  comment reads the same as one in a live command.
- **exportable** -- C, C++, Fortran, Rust, Python, JavaScript, and vignette
  chunks in those languages. Not read, but `export_unscanned()` writes them out
  for a scanner that reads them, blank-padded so that a finding's line number
  still points into the original file.
- **unexamined** -- present and accounted for, but not read: serialized `.rda`
  and `.rds` objects, binaries, and files no rule claims.

Coverage is never complete, and is not meant to be. What the `coverage` frame
offers is not completeness but legibility: a clean result can be checked rather
than trusted, because the scan says what it did not look at.

Every finding carries the R package lifecycle phases in which it runs --
`at_autoconf`, `at_build`, `at_check`, `at_install_src`, `at_install_bin`,
`at_load`, `at_attach`, `at_unload`, `at_detach` -- so code that executes on
`library()` is distinguishable from code that runs only when someone calls it.

A finding is not an accusation. Flagged files and code are often legitimate:
`configure` scripts exist for system-dependent configuration, and many packages
call system tools or download files for good reason. pkgaudit identifies what
deserves a reader's attention, not what is malicious.

For why this matters, see
[R Package Security](articles/r-package-security.html).
For the rule set, see
[Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

`untrustedpkg` has an `.onLoad()` hook containing a `system()` call, which runs
whenever the package is loaded; an ordinary function containing
`download.file()`, which runs at no phase because it runs only when someone
calls it; and a `configure` script containing `curl`, which runs when the
package is built, checked, or installed from source. Code that runs without
being asked deserves closer attention.

Phases overlap -- building a package with vignettes also installs and loads it
-- so one occurrence is counted under every phase it runs in.

`path = FALSE` omits the local filesystem path from output that will be shared.

Two functions carry a scan into other tools. `emit_sarif()` renders the result
as SARIF 2.1.0, which editors and code-scanning platforms read directly.
`export_unscanned()` writes the code pkgaudit cannot read into a directory for
a scanner that can. Both are covered in
[Getting Started with pkgaudit](articles/pkgaudit.html).

## Database integrity

pkgaudit detects security-relevant files and code using a SQLite database of
rules shipped at `inst/db/rules.db`. `load_rules()` verifies the database
against its bundled `.sha256` sidecar on every call and refuses to load a
modified one. To check an installed copy against the value published here:

``` r
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

Expected SHA-256: `r db_hash_md`

## Security

pkgaudit's own security model, and how to report a vulnerability in it, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/master/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/master/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
