---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit index.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# pkgaudit

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/test-coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
flags security-relevant files and code in R source packages for human review
without executing anything it scans. pkgaudit also models package lifecycle
execution semantics -- it reports not just what a package does, but when it
runs, so code that executes on install or load is distinguishable from code that
runs only when called.

A general-purpose scanner can read R -- Semgrep supports it experimentally --
but it reads files, not packages. Nothing tells it to look for R inside a help
file's `\examples{}` block or an `\Sexpr{}` macro, or inside a Sweave or R.rsp
vignette; and nothing tells it that `.onLoad()` runs on `library()`, that
`\Sexpr{}` evaluates while a help page is built, or that `configure` runs under
`R CMD check`. pkgaudit extracts R from wherever a package carries it, parses
the code for security-relevant patterns, and reports the lifecycle phases in
which each finding runs.

For why this matters, see [R Package Security](articles/security.html). For the
rule set, see [Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Four of `untrustedpkg`'s five findings run automatically:

* `.onLoad()` in `R/zzz.R` calls `system()`, which runs when users call
`library()`, and at build, check, and source installation, each of which loads
the package.

* An `\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help
page is rendered at build, check, and source installation.

* The `\examples{}` block in the same help file calls `download.file()` at
check.

* The `configure` script may invoke `curl` at build, check, and source
installation.

By contrast, `download.file()` in `R/fetch.R` sits in a regular function body,
so it is not known to run automatically and is reported under `none`.

Phases overlap -- building a package with vignettes also installs and loads it
-- so one occurrence is counted under every phase it runs in.

pkgaudit provides functions to integrate its scan with other tools.
`emit_sarif()` renders its results as SARIF 2.1.0, which editors and
code-scanning platforms read directly. `export_unscanned()` writes out package
code in languages pkgaudit cannot read, into a directory for a scanner that
can. See [Getting Started with pkgaudit](articles/pkgaudit.html) for details.

## Security

pkgaudit's own security model, and how to report a vulnerability, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
