---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit index.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# pkgaudit

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
reports which parts of a package can execute, when they execute, and what they
do, so that an untrusted package can be reviewed before it is installed and
loaded. pkgaudit never executes the code it scans.

A general-purpose scanner can read R -- Semgrep supports it experimentally --
but it reads files, not packages. Nothing tells it to look for R inside a help
file's `\examples{}` block or an `\Sexpr{}` macro, or inside a Sweave or R.rsp
vignette; and nothing tells it that `.onLoad()` runs on `library()`, that
`\Sexpr{}` evaluates while a help page is built, or that `configure` runs under
`R CMD check`. pkgaudit does both: it extracts R from wherever a package carries
it, and reports the lifecycle phases in which each finding runs.

A finding is not an accusation. `configure` scripts and calls to system tools,
for example, are often legitimate. pkgaudit helps to identify what deserves
reviewer attention, not what is malicious.

For why this matters, see [R Package Security](articles/security.html). For the
rule set, see [Rule Coverage](articles/rules.html).

## Installation

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

## Usage

A source package tarball can be scanned before it is installed. The example
below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration.

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

result <- audit_tarball(tarball)

summary(result, path = FALSE)
```

Four of `untrustedpkg`'s five findings run without anyone asking for them.
`.onLoad()` in `R/zzz.R` calls `system()`, so it runs on `library()` -- and at
build, check, and source installation, each of which loads the package. An
`\Sexpr{}` macro in `man/fetch_data.Rd` calls `httr::POST()` when the help page
is rendered, at those same three phases, and the `\examples{}` block in the same
file calls `download.file()`, which `R CMD check` runs. The `configure` script
may invoke `curl` at build, check, and source installation. The fifth finding is
the contrast that makes the rest legible: `download.file()` in `R/fetch.R` sits
in an ordinary function body, so it runs at no phase at all, reported under
`none`, because it executes only if someone calls it.

Two functions carry a scan into other tools. `emit_sarif()` renders the result
as SARIF 2.1.0, which editors and code-scanning platforms read directly.
`export_unscanned()` writes the code pkgaudit cannot read into a directory for
a scanner that can. See [Getting Started with pkgaudit](articles/pkgaudit.html)
for details.

## Security

pkgaudit's own security model, and how to report a vulnerability in it, are in
[SECURITY.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/SECURITY.md).
To propose or revise a rule, see
[CONTRIBUTING.md](https://github.com/tylerjssmith/pkgaudit/blob/main/.github/CONTRIBUTING.md).
How pkgaudit works internally, for a reader auditing the source, is in
[Internals](articles/internals.html).
