---
output:
  github_document:
    html_preview: false
---

<!-- index.md is generated from index.Rmd. Please edit this file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)

# Published SHA-256 of the shipped rules database. Read from the repository copy
# so the value below cannot drift from the database actually committed. knitr
# makes this file's own directory the working directory, so the repository root
# is one level up; knitting from the root is tolerated as well.
db_path <- file.path("..", "inst", "db", "rules.db")
if (!file.exists(db_path)) db_path <- file.path("inst", "db", "rules.db")
stopifnot(file.exists(db_path))
db_hash <- digest::digest(db_path, algo = "sha256", file = TRUE)

# Wrap in backticks for inline use. The backtick is built from its code point so
# that it does not terminate the inline expression that emits it.
backtick    <- intToUtf8(96L)
db_hash_md  <- paste0(backtick, db_hash, backtick)
```

# pkgaudit

[![R-CMD-check](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/R-CMD-check.yaml)
[![test coverage](https://raw.githubusercontent.com/tylerjssmith/pkgaudit/badges/coverage.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/coverage.yaml)
[![osv-scanner](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml/badge.svg)](https://github.com/tylerjssmith/pkgaudit/actions/workflows/osv-scanner.yaml)

pkgaudit is a static analysis security testing (SAST) tool for R packages. It
scans R source packages for files that can execute arbitrary commands during
autoconf, builds, checks, and installations, and for lifecycle hooks whose
bodies run automatically when a namespace is loaded, attached, unloaded, or
detached. It also scans R source code for security-relevant patterns like
`system()` calls.

R packages are the primary mechanism for sharing R code. They are also
[potential attack vectors](articles/r-package-security.html). pkgaudit helps you
identify which parts of an untrusted package deserve your attention before you
install and load it.

## Installation

You can install pkgaudit as follows:

``` r
remotes::install_github("tylerjssmith/pkgaudit")
```

pkgaudit detects security-relevant files and code using a SQLite database of
rules shipped with the package at `inst/db/rules.db`. To verify that the
installed copy of the database has not been modified since publication, check
its SHA-256 hash.

``` r
digest::digest(
  system.file("db", "rules.db", package = "pkgaudit"),
  algo = "sha256",
  file = TRUE
)
```

Expected SHA-256: `r db_hash_md`

## Usage

The [Getting Started with pkgaudit](articles/pkgaudit.html) and
[How pkgaudit Works](articles/how-it-works.html) vignettes document usage. The
[Rule Coverage](articles/rules.html) vignette documents the full rule set. The
example below scans `untrustedpkg`, a small package shipped with pkgaudit for
demonstration:

```{r usage}
library(pkgaudit)

tarball <- system.file(
  "extdata", "untrustedpkg", "untrustedpkg_0.1.0.tar.gz",
  package = "pkgaudit"
)

rules  <- load_rules()
result <- audit_tarball(tarball, rules = rules)
```

`summary.pkgaudit()` reports the file and code contexts found in untrustedpkg,
and counts the patterns found in each of them by the R package lifecycle phase
in which the code may execute (e.g., build, installation from source,
installation from binary, load). These phases can overlap (e.g., builds test
that a package can be installed from source and loaded), so a pattern may be
counted under more than one phase.

Below, we see that untrustedpkg includes a `configure` script, which is used
for system-dependent configuration but can execute arbitrary shell commands. It
also calls `system()` inside `.onLoad()`, which runs on `library(untrustedpkg)`,
and `download.file()` inside an ordinary function, which runs only if a user
calls the parent function and so belongs to no phase. Code that runs without
being asked deserves closer attention.

```{r usage-summary}
summary(result, path = FALSE)
```
