---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "", collapse = TRUE)
```

# Mapping an R Package's Execution Surface

The purpose of this project is to determine when code (R, Bash) in an R package
executes.

The method is direct measurement rather than reading the manual. `foo/` is an
instrumented probe package: every place a package can carry executable code
holds a marker that appends one line to a log when it runs. `run_foo.sh` runs a
single command with that log named, and `drive_foo.sh` runs every command in
`experiments.csv` in turn. A site that fires during `R CMD INSTALL` runs at
install; a site that never fires anywhere runs at no lifecycle phase at all.

Every marker line is `<time>\t<site>\t<pid>`. The pid matters because
`R CMD INSTALL` and `R CMD build` do parts of their work in subprocesses, and a
site firing in a subprocess is still a site that fired.

## Running the experiments

```sh
./drive_foo.sh experiments.csv      # everything
./run_foo.sh -e "R CMD INSTALL foo/" -o logs/install_src.log   # one command
```

Logs are written to `logs/`. Build artifacts are confined to `build/`, where
`foo` is a symlink to the package: `R CMD build`, `R CMD INSTALL --build` and
`R CMD check` all write to the working directory and offer no flag to redirect
it, so the working directory is what moves. Packages install into `lib/` rather
than the real user library.

`run_foo.sh` points `R_LIBS_USER` at `lib/`, which *replaces* the user library
rather than adding to it. The testing frameworks the probe exercises —
`testthat`, `tinytest` and `RUnit` — therefore have to be installed there too,
or their rows measure nothing. `lib/` is not tracked: when it is absent,
`run_foo.sh` rebuilds it by running `make_lib.R`, which installs every package
in `lib.lock` from CRAN and reports any version that differs from the lock,
the record of the library the shipped logs were measured against.

## The probe package

```{r files, echo = FALSE}
f <- list.files("foo", recursive = TRUE, all.files = TRUE, no.. = TRUE)
cat(paste0("foo/", sort(f[basename(f) != ".DS_Store"])), sep = "\n")
```

Files are named `probe.*`, or `probe_*.*` where more than one sits in the same
directory, **except** where R expects a particular name: `DESCRIPTION`,
`NAMESPACE`, `configure`, `cleanup`, `.Rprofile`, `src/Makevars`,
`src/install.libs.R`, `demo/00Index` and `inst/CITATION`.

Two naming details are load-bearing:

- `man/unix/` and `man/windows/` are merged into `man/` at install time, so a
  shared basename silently overwrites the base page. The first run of this
  experiment lost `man/probe_examples.Rd` entirely to `man/unix/` before the
  files were given distinct names.
- `inst/probe.R` is a control. `inst/` is copied verbatim into the installed
  package, so a plain R script there should never be sourced; it exists to show
  that whatever `inst/CITATION` does is specific to `CITATION`.

## Experiments

```{r experiments, echo = FALSE}
e <- read.csv("experiments.csv", stringsAsFactors = FALSE)
knitr::kable(e, col.names = c("Command (`-e`)", "Log (`-o`)"))
```

Row order matters. `R CMD build` has to run before anything that installs or
checks the tarball it produces, and `R CMD INSTALL --build` before the binary is
installed. `drive_foo.sh` clears `build/` first, so a command that fails before
writing its tarball cannot leave the previous one behind for later rows to test
silently.

## Results

```{r drive, echo = FALSE, cache = FALSE}
out <- system2("./drive_foo.sh", "experiments.csv", stdout = TRUE, stderr = TRUE)
cat(out, sep = "\n")
```

Read as a site-by-command matrix, with lifecycle commands first and the
user-invoked controls after:

```{r matrix, echo = FALSE}
sites_in <- function(path) {
  x <- readLines(path, warn = FALSE)
  x <- x[!startsWith(x, "#")]
  if (!length(x)) return(character(0))
  unique(vapply(strsplit(x, "\t", fixed = TRUE),
                function(p) if (length(p) >= 2L) p[[2L]] else NA_character_, ""))
}

order_from_csv <- basename(read.csv("experiments.csv", stringsAsFactors = FALSE)$o)
hits <- lapply(file.path("logs", order_from_csv), sites_in)
names(hits) <- sub("\\.log$", "", order_from_csv)

# Sites in the order they are first seen, so related ones stay together.
all_sites <- unique(unlist(hits, use.names = FALSE))

as_table <- function(cols) {
  m <- vapply(hits[cols], function(h) ifelse(all_sites %in% h, "x", ""), character(length(all_sites)))
  knitr::kable(data.frame(site = all_sites, m, check.names = FALSE, row.names = NULL))
}

lifecycle <- c("build", "build_novignette", "install_src", "install_tarball",
               "check", "check_ascran", "binary_make", "install_bin", "load",
               "attach", "detach", "unload")
as_table(lifecycle)
```

```{r matrix_controls, echo = FALSE}
as_table(setdiff(names(hits), lifecycle))
```

## What the results establish

**A function body runs when something calls it, and not otherwise — in every
context.** Each file context carries a pair of sites: `*_fn_called`, a function
invoked by top-level code in the same file, and `*_fn_uncalled`, one only
defined. Across all nineteen pairs, `*_fn_called` fires in exactly the set of
commands its context's top-level site fires in, and no `*_fn_uncalled` site
fires anywhere.

So there is nothing special about a function body's *position*. Whether it runs
is decided by whether it is reached, and the two bounds — reached from top
level, never reached — are the widest and narrowest readings available for code
whose call graph has not been traced. Both are measurements here rather than
inferences, which is what lets a scanner report one of them and say which.

**A framework can supply the caller.** `runit_fn_framework_called` sits inside
`test.probe()`, a function `inst/unitTests/runit.probe.R` defines and never
calls — and it fires during `R CMD check`, because RUnit calls it by name.
The uncalled bound is therefore not a floor in a test directory: a function that
looks unreached from inside the file may be reached from outside it.

**A lifecycle hook outside `R/` is not a hook.** `.onLoad` defined in
`data/probe.R` and in `tests/probe.R` never fires, in any command. Only code
that becomes the package namespace can supply one; elsewhere the same definition
ships as an ordinary object that nothing calls.

**The three testing layouts behave alike.** `tests/testthat/` — both a
`helper-*.R` file and a `test-*.R` file — `inst/tinytest/` and
`inst/unitTests/` all fire during `R CMD check` and nowhere else, matching plain
`tests/*.R`. Code inside a `test_that()` block fires with the rest of its file,
as expected: the block is a braced argument, not a function definition.

**Binary installation executes nothing.** `install_bin` is the only command that
fires no site at all.

**`src/Makevars` is the only unconditional build-time site.** It fires during
`R CMD build` even with `--no-build-vignettes`, because build runs make over
`src/` to clean object files before packaging, and a `:=` assignment is
evaluated while make parses the file. No compilation is involved.

**Everything else at build time is conditional on there being something to
evaluate.** With neither vignettes nor `\Sexpr`, `R CMD build` reaches only
`makevars` and `cleanup`. Add a vignette, or a help page containing `\Sexpr`,
and build installs and loads the package, reaching `configure`, `install_libs`,
`top_level`, `.onLoad` and the rest. Since both are common, the conservative
reading is that these contexts do run at build.

**A failed build is not a safety property.** Before TeX was available here, the
vignette step failed and `R CMD build` exited non-zero — after firing every one
of those sites. Code runs before the failure.

**`data/*.R` does not survive `R CMD build`.** It is evaluated at build time and
replaced by `data/probe.rda` in the tarball. It fires when installing from a
source *directory* and never when installing from the tarball, because by then
the R file is gone.

**`\Sexpr` stages differ, and no stage means install.** The unlabelled `\Sexpr`
is identical to `stage=install` across every experiment, as Writing R Extensions
says. `stage=render` does not run at either install; `stage=build` does not run
when installing a tarball, its result having been frozen into the Rd at build
time — the same evaluate-and-freeze pattern as `data/*.R`.

**`\examples` is check-only,** and the wrappers inside it divide three ways.
`\dontshow` and `\testonly` fire wherever the unwrapped code does. `\donttest`
is skipped by a plain `R CMD check` but fires under `--as-cran`, the check CRAN
performs, so it is not code that goes unrun. `\dontrun` fires nowhere.

**`inst/CITATION` runs during `R CMD check`,** and on `citation()`. The control,
`inst/probe.R`, never fires anywhere, so this is specific to the `CITATION`
filename and not a property of `inst/`.

**`demo/`, `tools/` and `exec/` run at no lifecycle phase.** Each fires only
under direct invocation, `exec/` for both the shell script and the R file. The
controls matter: without them, a silent log cannot distinguish "never reached"
from "instrumentation broken".

**`.Rprofile` at the package root fires** during `R CMD build` and installation
from a source directory — R picks it up when a process starts with its working
directory inside the package.

**`R/unix/` behaves exactly like `R/`; `R/windows/` never fires.** As expected on
a Unix-alike.

**`man/windows/` is not inert on Unix.** `rd_windows_sexpr_nostage` fires during
`R CMD build` and `R CMD check` even though the page is never installed here:
both process every Rd file under `man/`, including platform subdirectories.
`man/windows/`'s `\examples` never fires, because examples only run for
installed pages.

## `.Last.lib` and `.onDetach` are mutually exclusive

`.Last.lib` never appears in these logs, and that is the finding rather than a
gap. It is defined in `R/probe.R` alongside `.onDetach`, is exported, and is
present in both the package environment and the namespace — and `detach()`
still does not call it.

An isolated two-package experiment settles it:

| package defines | fires on `detach()` |
|---|---|
| `.Last.lib` only | `.Last.lib` |
| `.Last.lib` and `.onDetach` | `.onDetach` only |

So the two occupy the same phase and the same trigger — detaching an attached
package — but R calls only one, and `.onDetach` wins. `.Last.lib` is an
alternative, not a duplicate, which means a scan that looks only for `.onDetach`
misses the detach-time execution site of any package that defines just
`.Last.lib`.

## What was deliberately not tested

**`autoconf` / the `at_autoconf` phase.** R never runs `autoconf`; regenerating
`configure` from `configure.ac` is something a maintainer does, not something
that happens to a package a user installs. Probing it would measure a different
kind of event from every other phase here.

**`src/Makefile`.** R uses `src/Makefile` *instead of* R's default make process,
so it cannot coexist with `src/Makevars` in one package and would need a second
probe package. Assumed to execute at the same phases as `src/Makevars`: both are
make fragments R invokes at the same points in build, check and installation
from source.

**Windows variants** — `configure.win`, `configure.ucrt`, `cleanup.win`,
`cleanup.ucrt`, `Makevars.win`, `Makevars.ucrt`, `Makefile.win`,
`Makefile.ucrt`. These cannot be reached from a Unix-alike. Each is assumed to
execute at the same phases as its Unix counterpart, being the same mechanism
selected by platform. `R/windows/` and `man/windows/` *are* included above,
because their directories exist in a package built anywhere and their behavior
on Unix is itself worth knowing.

**`R_init_foo()`**, the load-time site belonging to compiled code. `src/probe.c`
exists only so that R runs make; it is not instrumented, because compiling is
not executing.

**`po/`.** Message catalogues are translation data, not code. There is no
evaluation step to instrument.
